---
title: "Matrix Algebra for Animal Breeders"
---

```{r}
#| label: ch01-setup
#| include: false
library(Matrix)
options(digits = 7)
```

::: {.callout-note icon=false}
## Learning Objectives

By the end of this chapter, you will be able to:

1. **Perform** transpose, addition, multiplication, and inversion **by hand** on matrices up to
   4 × 4, and **verify** each result in R
2. **Diagnose** a singular matrix, **explain** why breeding models routinely produce one, and
   **use** a generalized inverse
3. **Apply** the Kronecker and Hadamard products and **name** the chapter where each is used
4. **Decompose** a covariance matrix by Cholesky and by eigenvalues, and **state** what each
   decomposition is for
5. **Derive** the variance of a linear combination, $\mathrm{Var}(\mathbf{Ay}) =
   \mathbf{A}\,\mathrm{Var}(\mathbf{y})\,\mathbf{A}'$, and **use** it to assemble
   $\mathbf{V} = \mathbf{ZGZ}' + \mathbf{R}$
:::

*No prerequisites — this is the entry point.*

## Why linear algebra?

@fig-matrix-overview is this chapter on one page. Nothing in it needs to make sense yet — come
back to it when you have finished, and use it as a map in the meantime.

![Matrices in animal breeding. Panel 1 is §1.2, panel 2 is §1.6, panel 3 is §1.3 to §1.5, and
panel 4 is the applications map in §1.1.2 below.](../assets/figs/ch01-matrices-overview.png){#fig-matrix-overview
fig-alt="A four-panel infographic titled Matrices in Animal Breeding. Panel 1, what is a matrix: a
three by three array of the numbers 1 to 9, with definitions of element, row, column and dimension.
Panel 2, common matrix types: a vector of phenotypes, a square identity matrix, a symmetric
relationship matrix, and a diagonal residual variance matrix. Panel 3, basic operations: addition
and subtraction, scalar multiplication, matrix multiplication, and transpose. Panel 4, where
matrices are used: pedigree relationship matrices, genomic relationship matrices, the mixed model
equations, genetic evaluation and selection, and multiple-trait models."}

Two entries in the bottom row — the **genomic relationship matrix** and **single-step GBLUP** —
are outside this book's scope and belong to the companion volume on genomics. They are in the
figure so you can see where the pedigree-based methods you are about to learn eventually lead, not
because they are taught here.

### What it is

**Linear algebra is the mathematics of many quantities at once.** Ordinary algebra handles one
unknown at a time: solve $4x = 12$ and you have $x = 3$. Animal breeding never asks a question
that small. A genetic evaluation asks for the breeding values of every animal in a population
*simultaneously*, because no animal's value can be worked out without its relatives' — and those
depend back on it. That is not one equation. It is a hundred thousand equations that must all be
true at the same time.

Linear algebra is the language for writing down such a system and solving it. It gives you two
things ordinary algebra cannot:

1. **A way of writing that does not grow with the problem.** The same four symbols describe four
   animals or four million. You will see this happen in §1.7.
2. **Operations on whole collections of numbers at once** — an arithmetic where the objects being
   added and multiplied are entire tables rather than single values.

The word *linear* is a real restriction and worth knowing about. It means every unknown appears
only to the first power, multiplied by a constant and added to the others — no squares, no
products of two unknowns, no unknowns inside a logarithm. Most of quantitative genetics is built
to stay inside that restriction, because linear systems are the ones we can actually solve at
scale. The threshold models of Chapter 14 are what it looks like when a trait refuses to fit, and
what has to be done about it.

### Where it appears in animal breeding

Almost every quantitative method a breeder uses is linear algebra with a different name on it.
The table below is the map: the left column is where you meet the problem, the middle column is
what the algebra actually has to do, and the right column is the section of this chapter that
teaches it.

| Where it shows up | What the linear algebra does | Taught in |
|---|---|---|
| **The animal model** (Ch 6) | Build the mixed model equations [@henderson1975best], solve them for every effect at once, and read accuracy off the inverted coefficient matrix | §1.5, §1.9, §1.15 |
| **Relationship matrices** (Ch 4) | Build **A** from a pedigree and invert it — the object that makes the animal model practical [@henderson1976simple] | §1.9, §1.12 |
| **Selection index** (Ch 6) | Find the weights that combine an animal's own record with its relatives' by solving a linear system, $\mathbf{b} = \mathbf{P}^{-1}\mathbf{Ga}$ [@hazel1943genetic] | §1.9 |
| **Fixed effects and contemporary groups** (Ch 2, Ch 5) | Decide which effects can be estimated at all, and which herds can be compared with which | §1.8, §1.10 |
| **Multiple traits** (Ch 11) | Expand a small trait covariance matrix over every animal in the pedigree | §1.11 |
| **Random regression** (Ch 12) | Fit a curve per animal from a polynomial basis, and reduce twenty traits to the few dimensions carrying the variation | §1.13 |
| **Variance components** (Ch 20–22) | Build a likelihood out of quadratic forms, traces, and log-determinants, then differentiate it [@patterson1971recovery] | §1.12, §1.14 |
| **Simulation** (Ch 11, Ch 22) | Turn independent random draws into correlated breeding values with the right covariance structure | §1.12, §1.15 |
| **Solving at scale** (Ch 24) | Get an answer when the coefficient matrix has a million rows and cannot be inverted directly | §1.9 |
| **Optimal contribution selection** | Maximise genetic merit subject to a ceiling on average relationship — a quadratic form $\mathbf{x}'\mathbf{Ax}$ in the contributions | §1.14 · *theory in the companion volume on breeding programme design* |
| **Genomic prediction** | Replace **A** with a relationship matrix built from markers. The surrounding algebra does not change | *companion volume on genomics* |

Two things are worth noticing about that list before you start.

**The same few operations keep reappearing.** Multiply, transpose, invert, decompose. Optimal
contribution selection and REML look nothing like each other on the page, yet both rest on a
quadratic form — the object of §1.14 — and both need the same relationship matrix. Learning an
operation once buys you every application of it, which is why this chapter is organised by
operation rather than by application.

**The last two rows are deliberately out of scope here.** Optimal contribution selection belongs
to breeding programme design and genomic prediction to genomics; each has a companion volume, and
neither is taught in this book. They are in the table because the algebra underneath them is the
algebra in this chapter — a student who works through Chapter 1 is equipped for both.
### Why not just run the software?

Every operation in this chapter is one line of R, and nobody computes a national evaluation by
hand. So why learn to?

Because much of what goes wrong in a real evaluation is visible in the algebra before it is
visible in the results — and sometimes never visible in the results at all. An effect that cannot
be estimated still returns a number (§1.10). A genetic correlation of exactly 1 stops REML
converging, and the error names a matrix rather than the pair of traits that caused it (§1.13).
Two herds with no animals in common produce breeding values that look comparable and are not
(Chapter 5). In each case the software is behaving correctly; the person reading the output is the
one who has to recognise what happened.

The narrower reason is that Chapter 6 is unreadable without this material. Henderson's equations
are written in matrices, solved with a matrix inverse, and their accuracies read off a matrix
diagonal. A student who cannot form $\mathbf{X}'\mathbf{X}$ cannot see what those equations are
doing, and is left running software and trusting the output — which is exactly the position the
paragraph above says you should not be in.

It is worth knowing how far this transfers, too. Regression, ANOVA, and the general linear model
are the same computation as §1.9 with different words attached: `lm()` in R forms
$\mathbf{X}'\mathbf{X}$ and solves it, using the drop-a-level generalized inverse of §1.10.


### How to use this chapter

This is a reference. Read it once to see what is here, then come back to whichever section you
need — §1.16 maps every section to the chapters that consume it. Every section opens with real
numbers and works them before stating any rule, so landing in §1.14 without having read §1.4 still
works.

Three small examples carry the whole chapter, and they are reused on purpose: the algebra changes
while the animals stay the same, which is the only way to see that it is the algebra doing the
work.

## Matrices, vectors, and scalars

Before any operation, the objects themselves. Everything in this chapter is one of three things,
and a great deal of confusion later comes from losing track of which.

### The objects

Four lambs were measured for two traits — yearling weight in kilograms and loin muscle depth in
millimetres:

| Lamb | Yearling weight (kg) | Loin depth (mm) |
|---|---|---|
| 1 | 48 | 25 |
| 2 | 52 | 28 |
| 3 | 41 | 22 |
| 4 | 45 | 25 |

Strip away the headings and what is left is eight numbers in a rectangle, four rows deep and two
columns wide:

$$
\mathbf{Y} = \begin{bmatrix} 48 & 25 \\ 52 & 28 \\ 41 & 22 \\ 45 & 25 \end{bmatrix}
$$

That is a **matrix**, and it is the only object this chapter is really about. Nothing has been
modelled, assumed, or estimated. The rectangle is just the data, arranged so that position carries
meaning: row 3 is lamb 3, column 2 is loin depth, and the number where they meet — 22 — is lamb
3's loin depth.

```{r}
#| label: ch01-data-matrix
# The measurement table above, as a matrix. byrow = TRUE fills it the way the
# table reads: across row 1, then across row 2.
Y <- matrix(c(48, 25,
              52, 28,
              41, 22,
              45, 25), nrow = 4, byrow = TRUE)
Y

dim(Y)      # rows first, always: 4 by 2
Y[3, 2]     # row 3, column 2 -- lamb 3's loin depth, 22

# One column on its own is a vector. We will use the weights repeatedly, so
# give them a name now.
y <- Y[, 1]
y
```

::: {.callout-note .definition}
## Definition — Matrix, vector, and scalar

A **matrix** is a rectangular array of numbers in which position carries meaning. A **vector** is
a matrix with a single column. A **scalar** is a single number.

Its **dimensions** are quoted rows first: $\mathbf{Y}$ above is 4 × 2, read "four by two". The
number in row $i$ and column $j$ is called an **element** and written $y_{ij}$, again row first,
so $y_{32} = 22$.

**Why it matters.** These three words are the whole vocabulary of the chapter, and the book uses
typography to keep them apart: **bold uppercase** for a matrix ($\mathbf{Y}$, $\mathbf{X}$,
$\mathbf{A}$), **bold lowercase** for a vector ($\mathbf{y}$, $\mathbf{b}$), plain italic for a
scalar ($\sigma^2_e$, $\alpha$). When you meet $\mathbf{Y}$ and $\mathbf{y}$ in the same
equation in Chapter 11, they are different objects and the case is the only thing telling you so.
:::

Take one column of $\mathbf{Y}$ on its own — the four weights — and you have a vector, 4 × 1:

$$
\begin{bmatrix} 48 \\ 52 \\ 41 \\ 45 \end{bmatrix}
$$

This is the shape almost everything in the book arrives in: one row per record, in a fixed order
that never changes. That fixed order is what lets a second table — which flock each lamb is in,
which sire it is by — line up against it row by row without anyone having to say so.

### The one idea underneath all of it

Suppose you wanted a single index combining both traits, counting a kilogram of weight as worth 1
and a millimetre of depth as worth 2. For lamb 1 that is

$$
1(48) + 2(25) = 98
$$

Multiply each quantity by a weighting, add up the results. That operation — and *only* that
operation — is what linear algebra automates.

::: {.callout-note .definition}
## Definition — Linear combination

A **linear combination** of a set of quantities is what you get by multiplying each one by a
constant and adding the results: $c_1x_1 + c_2x_2 + \cdots + c_nx_n$. The constants are called
**coefficients** or **weights**.

**Why it matters.** This is the single idea the whole chapter is built from, and recognising it
saves you learning the same thing five times. Matrix multiplication (§1.5) is linear combinations
performed wholesale. A fitted value $\mathbf{Xb}$ is a linear combination of effects. A selection
index is a linear combination of trait records. A breeding value is a linear combination of the
information on an animal and its relatives. Every one of them is "multiply by weights, then add" —
what changes between them is only *where the weights come from*.
:::

The same weights applied to every lamb at once, still one multiplication and one addition per
term:

```{r}
#| label: ch01-linear-combination
1 * Y[, 1] + 2 * Y[, 2]     # 98 for lamb 1, and the other three
```

Section 1.5 replaces that line with a single matrix product. Nothing about the arithmetic changes
— only how much of it you have to write down.

### Notation

These are conventions, not results. There is nothing to discover — they are names, and you should
be handed them rather than led to them.

The book writes **bold uppercase** for a matrix, **bold lowercase** for a vector, and plain italic
for a scalar, as set out above. Two more pieces of vocabulary before we start operating on them.

A matrix is **square** when it has as many rows as columns. Only square matrices have
determinants, inverses, and eigenvalues, so a good deal of this chapter applies to them alone. Our
$\mathbf{Y}$ is 4 × 2 and therefore not square — but $\mathbf{Y}'\mathbf{Y}$ would be, and it is
no accident that the products at the centre of every model in this book come out square however
many records you have.

The **main diagonal** of a square matrix runs from its top-left corner to its bottom-right: the
elements $a_{11}, a_{22}, a_{33}, \ldots$, the ones whose row and column index are the same.
Almost everything interesting about a square matrix is either on that diagonal or defined
relative to it.

## Transpose

The simplest operation there is. It needs nothing but the matrix itself, which is why it comes
first.

Write the measurement matrix $\mathbf{Y}$ from §1.2, and then write it again with its rows and
columns interchanged:

$$
\mathbf{Y} = \begin{bmatrix} 48 & 25 \\ 52 & 28 \\ 41 & 22 \\ 45 & 25 \end{bmatrix}
\qquad
\mathbf{Y}' = \begin{bmatrix} 48 & 52 & 41 & 45 \\ 25 & 28 & 22 & 25 \end{bmatrix}
$$

$\mathbf{Y}$ is 4 × 2 and $\mathbf{Y}'$ is 2 × 4. Row 1 of $\mathbf{Y}$ — lamb 1's two
measurements — has become column 1 of $\mathbf{Y}'$. The entry that sat in row 3, column 2 now
sits in row 2, column 3. That is the whole operation, and written out it is simply

$$
y'_{ij} = y_{ji}
$$

The transpose is written $\mathbf{Y}'$ in this book and read "Y prime"; some texts write
$\mathbf{Y}^{T}$ for the same thing. Transposing twice puts everything back, so
$(\mathbf{Y}')' = \mathbf{Y}$.

```{r}
#| label: ch01-transpose
t(Y)                    # 4 x 2 becomes 2 x 4
all.equal(t(t(Y)), Y)   # transposing twice returns the original
```

A square matrix that is unchanged by transposing is called **symmetric**: its element in row $i$,
column $j$ equals the one in row $j$, column $i$, so it is a mirror image of itself across the
diagonal running from top-left to bottom-right.

$$
\begin{bmatrix} 1 & 0.5 \\ 0.5 & 1 \end{bmatrix}
$$

That matters more than it looks. **Every relationship matrix and every covariance matrix in this
book is symmetric**, because the relationship between animal 1 and animal 2 is the same number as
the relationship between animal 2 and animal 1. So is $\mathbf{X}'\mathbf{X}$ (§1.7), and so is
the whole left-hand side of the mixed model equations in Chapter 6. It is a fact used constantly
and almost never announced, and it is worth carrying: whenever you see $\mathbf{A}'$ written for
a covariance matrix, it is the same matrix as $\mathbf{A}$.


## Addition, subtraction, and scalar multiplication

Like the transpose, these three leave every number where it is. Nothing travels; position is
preserved. That is exactly what makes them easy, and it is the property matrix multiplication
gives up in §1.5.

### Addition and subtraction

Matrix addition is element by element, position preserved. Here is the case worth caring about.

For sheep yearling weight (kg) and loin muscle depth (mm), suppose the additive genetic covariance
matrix and the residual covariance matrix are

$$
\mathbf{G} = \begin{bmatrix} 36 & 12 \\ 12 & 9 \end{bmatrix}
\qquad
\mathbf{R} = \begin{bmatrix} 64 & 4 \\ 4 & 16 \end{bmatrix}
$$

Call these **Example B**. They are the two-trait covariance structure the rest of the chapter
returns to.

Adding them position by position — 36 + 64, 12 + 4, 12 + 4, 9 + 16 — gives the phenotypic
covariance matrix:

$$
\mathbf{P} = \mathbf{G} + \mathbf{R} = \begin{bmatrix} 100 & 16 \\ 16 & 25 \end{bmatrix}
$$

Everything a breeder quotes comes off these three matrices by division. The phenotypic standard
deviations are $\sqrt{100} = 10$ kg and $\sqrt{25} = 5$ mm. The heritabilities are the diagonal of
$\mathbf{G}$ over the diagonal of $\mathbf{P}$: $36/100 = 0.36$ for weight and $9/25 = 0.36$ for
depth. The genetic correlation is the off-diagonal of $\mathbf{G}$ over the square root of the
product of its diagonals, $12/\sqrt{36 \times 9} = 12/18 = 0.67$; the residual correlation is
$4/\sqrt{64 \times 16} = 0.125$; the phenotypic correlation is $16/(10 \times 5) = 0.32$.

```{r}
#| label: ch01-add-covariance
G <- matrix(c(36, 12,
              12,  9), nrow = 2)   # additive genetic covariances
R <- matrix(c(64,  4,
               4, 16), nrow = 2)   # residual covariances
P <- G + R                         # phenotypic covariances, element by element
P

# Heritabilities: the diagonal of G over the diagonal of P. diag() pulls out
# the main diagonal of a square matrix -- the elements defined in section 1.2.
diag(G) / diag(P)

# Correlations: an off-diagonal over the root of the product of its diagonals.
G[1, 2] / sqrt(G[1, 1] * G[2, 2])   # genetic,     0.667
R[1, 2] / sqrt(R[1, 1] * R[2, 2])   # residual,    0.125
P[1, 2] / sqrt(P[1, 1] * P[2, 2])   # phenotypic,  0.32
```

Subtraction works the same way, and it is how $\mathbf{R}$ is usually obtained in practice:
$\mathbf{R} = \mathbf{P} - \mathbf{G}$. Addition requires identical dimensions — there is no
sensible answer to "what is the 2 × 2 plus the 3 × 3," and R will refuse.

**Scalar multiplication** multiplies every element by the same number. It is how the relationship
matrix becomes a covariance matrix: $\mathbf{G} = \mathbf{A}\sigma^2_a$ for a single trait, so for
two full sibs with $\sigma^2_a = 36$,

$$
\mathbf{A}\sigma^2_a = \begin{bmatrix} 1 & 0.5 \\ 0.5 & 1 \end{bmatrix} \times 36
= \begin{bmatrix} 36 & 18 \\ 18 & 36 \end{bmatrix}
$$

The variance ratio $\alpha = \sigma^2_e / \sigma^2_a$ is a scalar too, and it is the number that
controls how much a mixed model shrinks its predictions. With the values above,
$\alpha = 64/36 = 1.78$.

## Matrix multiplication

Transpose, addition and scalar multiplication all leave the numbers where they are. Multiplication
does not, and it is the operation everything else in the book is built on. It is also the one
students find hardest, so we are going to do it before naming it.

### Two totals

Flock A's two lambs weigh 48 and 52 kg; flock B's weigh 41 and 45. We want each flock's total
weight. Here are the records, with a bookkeeping column marking which flock each lamb belongs to.
Call that table $\mathbf{W}$. Nothing is being modelled or estimated — we are adding up weights,
and the ones and zeros are only there to say which lamb belongs in which total.

$$
\begin{array}{c|cc|c}
 & \text{flock A} & \text{flock B} & y \\ \hline
\text{lamb 1} & 1 & 0 & 48 \\
\text{lamb 2} & 1 & 0 & 52 \\
\text{lamb 3} & 0 & 1 & 41 \\
\text{lamb 4} & 0 & 1 & 45
\end{array}
$$

Flock A's total: walk down the flock A column, pair each entry with the weight beside it, multiply,
and add.

$$
1(48) + 1(52) + 0(41) + 0(45) = 100
$$

Flock B's total, the same walk down the other column:

$$
0(48) + 0(52) + 1(41) + 1(45) = 86
$$

Two totals, from two columns, against one list of weights. Write them as a pair:

$$
\begin{bmatrix} 100 \\ 86 \end{bmatrix}
$$

{{< animation ch01-matmul caption="Row times column, one lamb at a time. The inner dimensions cancel: a 2 × 4 against a 4 × 1 gives a 2 × 1." >}}

**What mattered and what did not.** The numbers 48 and 52 were these lambs' weights — change the
weights and the totals change. The number of columns we walked through was the number of flocks,
and it fixed how many totals came out: two columns in, two totals out. But *the walk itself* —
take a column, pair it term by term with the list, multiply, add — never changed, and it would not
change for 4 lambs or 4 million, for 2 flocks or 200. The one thing the walk needed was that the
column and the list be the same length, so the pairing never ran out on either side.

**That walk is matrix multiplication.** Stand the bookkeeping table on its side so its columns
become rows — that is the transpose from §1.3 — and the pair of totals is $\mathbf{W}'\mathbf{y}$,
whose $i$th element pairs row $i$ of $\mathbf{W}'$ with $\mathbf{y}$ term by term and sums:

$$
(\mathbf{W}'\mathbf{y})_i = \sum_k (\mathbf{W}')_{ik}\, y_k
$$

In general, for $\mathbf{A}$ of size $m \times n$ and $\mathbf{B}$ of size $n \times p$, the
product $\mathbf{C} = \mathbf{AB}$ is $m \times p$ with

$$
c_{ij} = \sum_{k=1}^{n} a_{ik} b_{kj}
$$

Element $c_{ij}$ is row $i$ of the left matrix walked against column $j$ of the right one. The
condition for this to work is exactly the one we relied on without saying so: **the inner
dimensions must match**. An $m \times n$ times an $n \times p$ works and gives $m \times p$; the
two $n$'s meet in the middle and cancel, and what survives is the outside pair.

::: {.callout-caution .check collapse="true"}
## Check Yourself

You just multiplied a 2 × 4 by a 4 × 1 and got a 2 × 1. Before reading on: what comes out of a
4 × 2 times a 4 × 1, and why?

**Answer.** Nothing — it cannot be formed. The inner dimensions are 2 and 4, and they do not
match, so the pairing runs out after two terms with two weights left over. Written the other way,
$\mathbf{W}'\mathbf{y}$ is 2 × 4 times 4 × 1 and works. This is why the bookkeeping table gets
transposed before it meets $\mathbf{y}$, and not for any deeper reason.
:::

::: {.callout-note .definition}
## Definition — Conformable

Two matrices are **conformable** for an operation when their dimensions permit it. For addition
and subtraction that means identical dimensions. For multiplication it means something weaker,
which §1.5 arrives at by doing it rather than by declaring it.

**Why it matters.** Non-conformability is the most common error a student makes in R, and it is
the useful kind: the operation stops rather than returning a wrong number. When `%*%` throws
"non-conformable arguments", it is telling you that the two objects do not line up — usually
because a matrix needs transposing, and occasionally because the model itself is wrong.
:::

The special matrices below are named by where they appear in this book, because that is how you
will meet them.

```{r}
#| label: ch01-matmul
# The bookkeeping table of the walk above: one column per flock.
W <- cbind(flockA = c(1, 1, 0, 0),
           flockB = c(0, 0, 1, 1))
W

# Transpose it, then multiply. The two flock totals we worked by hand: 100, 86.
t(W) %*% y

# W'W counts instead of summing: how many lambs are in each flock.
t(W) %*% W
```


::: {.callout-tip}
## In R — `%*%` is multiplication; `*` is not

`A * B` in R multiplies element by element and is the Hadamard product of §1.11. Matrix
multiplication is `A %*% B`. The trap is that for two matrices of the same shape, `A * B` runs
without error and returns something plausible-looking, so this fails silently.

Once you have written `t(X) %*% X` a few times, use `crossprod(X)` instead — and `crossprod(X, y)`
for $\mathbf{X}'\mathbf{y}$. They give the same answer and are faster, because they never build
$\mathbf{X}'$ at all. There is also `tcrossprod(X)` for $\mathbf{XX}'$. The rest of this book uses
the `crossprod` forms.

```{r}
#| label: ch01-crossprod
all.equal(crossprod(W), t(W) %*% W)
all.equal(crossprod(W, y), t(W) %*% y)
```
:::

### The rule that catches everyone

Now something you cannot read off the page. Take two 2 × 2 matrices with no symmetry and nothing
in common:

$$
\mathbf{A} = \begin{bmatrix} 2 & 1 \\ 0 & 3 \end{bmatrix}
\qquad
\mathbf{B} = \begin{bmatrix} 1 & 4 \\ 2 & 0 \end{bmatrix}
$$

Multiply them — you now know how — and transpose the result:


$$
\mathbf{AB} = \begin{bmatrix} 4 & 8 \\ 6 & 0 \end{bmatrix}
\qquad
(\mathbf{AB})' = \begin{bmatrix} 4 & 6 \\ 8 & 0 \end{bmatrix}
$$

There are two obvious candidates for what $(\mathbf{AB})'$ ought to equal. Form both:

$$
\mathbf{A}'\mathbf{B}' = \begin{bmatrix} 2 & 4 \\ 13 & 2 \end{bmatrix}
\qquad
\mathbf{B}'\mathbf{A}' = \begin{bmatrix} 4 & 6 \\ 8 & 0 \end{bmatrix}
$$

The second one matches. The first is not even close.

**What mattered and what did not.** The particular entries did not matter — nothing about 2, 1, 0
or 3 made the second candidate win. The dimensions did not matter either; had $\mathbf{A}$ been
3 × 5 and $\mathbf{B}$ 5 × 2, $\mathbf{A}'\mathbf{B}'$ would have been 5 × 3 times 2 × 5 and could
not have been formed at all. What mattered is that transposing swaps which index runs along which
side, so the two factors have to swap as well for the ends still to meet.

::: {.callout-important .key-equation}
## Key Equation — The reversal rules

$$
(\mathbf{AB})' = \mathbf{B}'\mathbf{A}'
$$ {#eq-reversal}

**In words:** transposing a product reverses the order of the factors. Watch for the shape of this
rule again later — the reversal is not something special about the transpose.
:::

```{r}
#| label: ch01-transpose-add
# matrix() fills COLUMN by column, so read these down the columns, not across.
A <- matrix(c(2, 0,
              1, 3), nrow = 2)
B <- matrix(c(1, 2,
              4, 0), nrow = 2)
A
B

t(A %*% B)      # (AB)'
t(B) %*% t(A)   # B'A' -- the same
t(A) %*% t(B)   # A'B' -- not the same
```

::: {.callout-warning}
## Common Mistake — assuming AB and BA are the same thing

They are usually not, and often only one of them can even be formed. With the same $\mathbf{A}$
and $\mathbf{B}$:

$$
\mathbf{AB} = \begin{bmatrix} 4 & 8 \\ 6 & 0 \end{bmatrix}
\qquad
\mathbf{BA} = \begin{bmatrix} 2 & 13 \\ 4 & 2 \end{bmatrix}
$$

Same two matrices, different products. Matrix multiplication is **associative**
($(\mathbf{AB})\mathbf{C} = \mathbf{A}(\mathbf{BC})$) and **distributive**
($\mathbf{A}(\mathbf{B}+\mathbf{C}) = \mathbf{AB} + \mathbf{AC}$), so you may re-bracket freely —
but you may never reorder. When a derivation in a later chapter moves a matrix across another one,
it is doing something, and it is worth checking what.
:::

This is the single most productive source of error in the chapters that follow. When you are
assembling the mixed model equations and you need the block above the diagonal, it is
$\mathbf{X}'\mathbf{Z}$, and the block below it is $\mathbf{Z}'\mathbf{X}$ — which is
$(\mathbf{X}'\mathbf{Z})'$, not $\mathbf{X}'\mathbf{Z}$ again. Having formed
$\mathbf{A}'\mathbf{B}'$ once by hand and watched it come out wrong is worth more than reading the
rule three times.

## Special matrices

Some matrices turn up so often that they have names. They are worth meeting *after* the
operations, because most of them are defined by how they behave under those operations rather
than by what they look like.

| Matrix | What it is | Where it shows up |
|---|---|---|
| Identity $\mathbf{I}$ | 1s on the diagonal, 0s elsewhere. Multiplying by it changes nothing: $\mathbf{AI} = \mathbf{IA} = \mathbf{A}$ | the residual structure $\mathbf{R} = \mathbf{I}\sigma^2_e$; the "no relationships" model of Ch 3 |
| Null $\mathbf{0}$ | all zeros | the off-diagonal blocks joining unrelated families |
| $\mathbf{J}$ | all ones | the mean model; $\mathbf{J}/n$ as the averaging operator |
| Diagonal | non-zero only on the diagonal | weights, unequal residual variances, $\mathbf{D}$ in $\mathbf{A} = \mathbf{TDT}'$ (Ch 4) |
| Symmetric | $a_{ij} = a_{ji}$ | every covariance matrix; $\mathbf{X}'\mathbf{X}$; the whole MME left-hand side |
| Triangular | zeros above (or below) the diagonal | $\mathbf{L}$ from Cholesky (§1.12); $\mathbf{T}$ from the pedigree (Ch 4) |
| Incidence | entries 0 or 1 only | $\mathbf{X}$ and $\mathbf{Z}$ — almost entirely zeros, which is why Ch 24 exists |
| Block / partitioned | built from smaller matrices as tiles | the MME itself (Ch 6) |
| Positive definite | a validity condition — see §1.13 | what a covariance matrix must be to be a covariance matrix |

```{r}
#| label: ch01-special
diag(3)                        # the identity, 3 x 3
diag(c(2, 5, 9))               # a diagonal matrix from a vector
matrix(1, nrow = 2, ncol = 3)  # J, all ones, 2 x 3

# Y is the measurement matrix from section 1.2.
nrow(Y)
ncol(Y)
t(Y)                           # not square: transposing turns 4 x 2 into 2 x 4

# A matrix that is mostly zeros, stored sparsely: only the positions of the
# non-zeros are kept, and a dot is printed wherever a zero would be.
Matrix(diag(4), sparse = TRUE)
```

::: {.callout-note .beyond}
## Beyond This Book — how X is really stored

The incidence matrices $\mathbf{X}$ and $\mathbf{Z}$ in the table above are the reason this
matters. The $\mathbf{X}$ built in §1.7 is 4 × 2 and almost half zeros. In a national evaluation
it is millions of rows
by tens of thousands of columns and more than 99.99% zeros, and no one ever writes it out. Sparse
storage — keeping only the positions of the non-zeros — is what makes such a matrix fit in memory
at all, and it changes how every operation in this chapter is actually performed. Chapter 24 shows
*why* dense methods fail at that scale; the mechanics of sparse storage and factorisation belong
to the companion volume on large-scale computation.
:::

One more piece of notation before we start: $\mathbf{X}'$ — read "X transpose", or more often
just "X prime" — is the subject of §1.3. Some texts write $\mathbf{X}^{T}$ for the same thing.

## Why matrices: one model, many animals

Four lambs were weighed at a year old. Two are from flock A, two from flock B.

| Lamb | Flock | Yearling weight (kg) |
|---|---|---|
| 1 | A | 48 |
| 2 | A | 52 |
| 3 | B | 41 |
| 4 | B | 45 |

: Example A — four lambs, two flocks. {#tbl-lambs}

We think each lamb's weight is a flock level plus whatever is left over for that individual.
Written one lamb at a time, that is four separate equations. Take flock B as the baseline and let
$b_1$ be flock B's level and $b_2$ be however much flock A runs above it:

$$
\begin{aligned}
48 &= b_1 + b_2 + e_1 \\
52 &= b_1 + b_2 + e_2 \\
41 &= b_1 \phantom{{}+ b_2} + e_3 \\
45 &= b_1 \phantom{{}+ b_2} + e_4
\end{aligned}
$$

Every equation has the same two unknowns. What differs between them is only *which* unknowns
appear — a 1 or a 0 in front of each. Pull those coefficients out and write them as a table, one
row per lamb, one column per unknown:

$$
\begin{bmatrix} 48 \\ 52 \\ 41 \\ 45 \end{bmatrix}
=
\begin{bmatrix} 1 & 1 \\ 1 & 1 \\ 1 & 0 \\ 1 & 0 \end{bmatrix}
\begin{bmatrix} b_1 \\ b_2 \end{bmatrix}
+
\begin{bmatrix} e_1 \\ e_2 \\ e_3 \\ e_4 \end{bmatrix}
$$

Four rows, one per record. Two columns, one per thing being estimated: the first column is all
ones because every lamb has flock B's baseline in it, and the second column is 1 exactly when the
lamb is in flock A. Call the list of weights $\mathbf{y}$, the table of ones and zeros
$\mathbf{X}$, the two unknowns $\mathbf{b}$, and the leftovers $\mathbf{e}$:

$$
\mathbf{y} = \mathbf{Xb} + \mathbf{e}
$$

<!-- {{< animation ch01-stacking caption="Four scalar equations stack into y = Xb + e." >}} -->

Nothing has been gained yet — those are the same four equations with the arithmetic hidden. What
has been gained is that the statement no longer mentions four. Add a fifth lamb and
$\mathbf{y}$ gets a fifth entry and $\mathbf{X}$ a fifth row; the equation is unchanged. Add
40,000 lambs and it is still unchanged. That invariance is the whole reason this book is written
in matrices: **the model statement stops depending on how many animals you have**, and by
Chapter 6 the same four symbols will carry 100,000 pigs without a character of difference.

```{r}
#| label: ch01-build-Xy
lambs <- data.frame(lamb  = 1:4,
                    flock = c("A", "A", "B", "B"),
                    wt    = c(48, 52, 41, 45))
lambs

# y is just the column of weights.
y <- lambs$wt
y

# X, built exactly as it is written on the page: a column of ones for flock B's
# baseline, then a 1 wherever the lamb is in flock A.
baseline <- c(1, 1, 1, 1)
flockA   <- c(1, 1, 0, 0)
X <- cbind(baseline, flockA)
X
```

With $\mathbf{X}$ and $\mathbf{y}$ in hand, the two products of §1.5 finally have their
book meanings:

```{r}
#| label: ch01-xtx
crossprod(X)        # X'X -- the table of counts
crossprod(X, y)     # X'y -- the table of sums

# Xb hands each record its fitted value. b is solved for in §1.9.
b_hat <- c(43, 7)
X %*% b_hat
```

R will also build $\mathbf{X}$ for you from the data frame. That is what you will use in
practice, but it is worth seeing only after you have built one by hand, because the function
chooses the parameterisation and you need to be able to check which one it chose.

```{r}
#| label: ch01-model-matrix
# The same X, from a formula. Listing flock B first makes it the baseline,
# which is the parameterisation we wrote out above.
lambs$flock <- factor(lambs$flock, levels = c("B", "A"))
X_auto <- model.matrix(~ flock, data = lambs)
X_auto

# Two cosmetic differences from the X we built by hand: the first column is
# called "(Intercept)" rather than "baseline", and model.matrix() attaches
# bookkeeping attributes of its own. Ignore both -- the numbers are identical.
all.equal(X, X_auto, check.attributes = FALSE)
```

### X′X counts and X′y sums

Now form $\mathbf{X}'\mathbf{X}$ for the $\mathbf{X}$ just built — the one whose first column is all ones and whose second
column flags flock A — and form $\mathbf{X}'\mathbf{X}$ from it. Row 1 of $\mathbf{X}'$ against column 1 of $\mathbf{X}$
is $1(1) + 1(1) + 1(1) + 1(1) = 4$. Row 1 against column 2 is $1(1) + 1(1) + 1(0) + 1(0) = 2$. Row
2 against column 2 is $1(1) + 1(1) + 0(0) + 0(0) = 2$.

$$
\mathbf{X}'\mathbf{X} = \begin{bmatrix} 4 & 2 \\ 2 & 2 \end{bmatrix}
\qquad
\mathbf{X}'\mathbf{y} = \begin{bmatrix} 186 \\ 100 \end{bmatrix}
$$

<!-- {{< animation ch01-xtx-counts caption="X'X counts records; X'y sums them." >}} -->

Look at what those four numbers are. Every element of $\mathbf{X}$ is a 0 or a 1, so multiplying
two of them together gives 1 only when both are 1 — and summing the products just *counts the rows
where both columns had a 1*.

- $\mathbf{X}'\mathbf{X}$ **is a table of counts.** The 4 is the number of records. The bottom-right 2 is the
  number of records in flock A. The off-diagonal 2 counts the records that have both
  a baseline and a flock A indicator — also 2, because every flock A record has both.
- $\mathbf{X}'\mathbf{y}$ **is a table of sums.** 186 is the grand total of all four weights; 100
  is the flock A total we computed above.

These are not facts stated about $\mathbf{X}'\mathbf{X}$ after the fact. They are the reason it
gets built. Every left-hand side you will meet in this book is a table of counts with variance
ratios added to some of its diagonal, and every right-hand side is a table of sums. A student who
owns this reads Chapter 6's mixed model equations without effort.

Two more products worth naming while we are here. $\mathbf{Xb}$ hands each record its fitted
value — with the solution we reach in §1.9, $\mathbf{b} = (43, 7)'$, it gives
$(50, 50, 43, 43)'$, flock A's lambs getting $43 + 7$ and flock B's getting 43. And $\mathbf{Zu}$, which arrives in Chapter 3, hands each
record its own animal's breeding value in exactly the same way.

## Rank, singularity, and the determinant

The parameterisation in §1.7 was a choice. A more natural-looking one gives every flock its own
column and keeps the baseline as well:

$$
\mathbf{X}_3 = \begin{bmatrix}
1 & 1 & 0 \\ 1 & 1 & 0 \\ 1 & 0 & 1 \\ 1 & 0 & 1
\end{bmatrix}
$$

Three columns: a baseline, a flock A indicator, a flock B indicator. Look at them. The second
column plus the third column equals the first, in every row. The third column carries no
information the other two lack — knowing a lamb is not in flock A already tells you it is in
flock B.

Columns that can be built from other columns are **linearly dependent**, and $\mathbf{X}_3$ has
three columns of which only two carry independent information. Build it and check the claim
directly — column 2 plus column 3 really is column 1:

```{r}
#| label: ch01-dependent-column
X3 <- cbind(baseline = 1,
            flockA   = c(1, 1, 0, 0),
            flockB   = c(0, 0, 1, 1))
X3

X3[, "flockA"] + X3[, "flockB"]      # equals the baseline column, every row
all.equal(X3[, "flockA"] + X3[, "flockB"], X3[, "baseline"],
          check.attributes = FALSE)
```

R will tell you the rank directly. `qr()` factorises the matrix and its `$rank` component counts
the independent columns:

```{r}
#| label: ch01-rank
qr(X3)$rank        # 2 -- not 3, even though there are three columns
ncol(X3)           # 3 columns, so X3 is rank deficient

qr(X)$rank         # the two-column X of section 1.7 is full rank
ncol(X)
```

::: {.callout-note .definition}
## Definition — Linear dependence and rank

A set of columns is **linearly dependent** when at least one of them can be written as a linear
combination (§1.1) of the others — it adds no information the rest do not already carry. Columns
that cannot be are **linearly independent**.

The **rank** of a matrix is the number of linearly independent columns it has. A matrix whose rank
equals its number of columns is **full rank**; one whose rank is smaller is **rank deficient**.
$\mathbf{X}_3$ has three columns and rank 2, so it is rank deficient.

**Why it matters.** Rank is what decides whether a model's effects can be estimated at all. In
animal breeding the answer is routinely no, and rank deficiency is normal rather than exceptional
— every model with a mean and a contemporary group effect has it. Chapter 2 is largely about what
to do next, and §1.10 gives the tool.
:::

Form $\mathbf{X}_3'\mathbf{X}_3$ and the dependence is inherited:

$$
\mathbf{X}_3'\mathbf{X}_3 = \begin{bmatrix}
4 & 2 & 2 \\ 2 & 2 & 0 \\ 2 & 0 & 2
\end{bmatrix}
$$

Again column 2 plus column 3 equals column 1 — the dependence has been inherited:

```{r}
#| label: ch01-xtx-singular
XtX3 <- crossprod(X3)
XtX3

XtX3[, 2] + XtX3[, 3]     # still equals column 1
qr(XtX3)$rank             # 2, in a 3 x 3 matrix
```

::: {.callout-note .definition}
## Definition — Singular matrix

A square matrix is **singular** when its columns are linearly dependent — equivalently, when its
rank is less than its dimension, and when its determinant is zero. Those three statements say the
same thing about the same matrix. A square matrix that is not singular is **non-singular**, or
**full rank**.

A fourth equivalent statement is that a singular matrix has no inverse; §1.9 defines the inverse
and shows why the two ideas are the same one.

The practical meaning is that the system of equations the matrix defines has **no unique
solution**. Here you could add any constant to the baseline and subtract it from both flock
effects without changing a single fitted value, so there is no single right answer to report.

**Why it matters.** This is the word your software will use when it refuses to fit a model, and it
is worth being able to translate it into "two of the things I asked for cannot be told apart from
each other". The response is not to give up: §1.10 solves the system anyway and tells you which
quantities survive. In Chapter 6 the animal block of the mixed model equations is
deliberately made non-singular by adding $\mathbf{A}^{-1}\alpha$ to it, which is why the animal
model can return a breeding value even for an animal with no record of its own.
:::

<!-- {{< animation ch01-singular caption="A dependent column collapses the determinant to zero." >}} -->

### Where this material comes from

Rank, linear dependence and the determinant are nineteenth-century mathematics with no single
originating paper, so nothing is cited for them above. What *does* have a literature is the
question this section leads into: what to do with a model whose $\mathbf{X}$ is not full rank.

Two results settle it, and both arrived surprisingly late. R. C. Bose gave the definition of an
**estimable function** in the 1940s. Penrose [@penrose1955generalized] defined the generalized
inverse for arbitrary matrices, and Rao [@rao1962note] showed what it does for least squares when
the normal equations are singular.

Those dates are worth pausing on, because they fall *after* Henderson had begun the work this book
teaches. Searle — who took his doctorate in animal breeding under Henderson at Cornell — put it
plainly in a later interview [@wells2009conversation]: his group had "not kept up with the concept
of estimability propounded by R. C. Bose", nor were they "aware of Penrose's (1955) generalized
inverse matrix which, as Rao (1962) demonstrated, clarified the whole business of solving least
squares equations which are so often not of full rank."

§1.10 is that clarification, and Chapter 2 makes estimability precise.

**This is the normal case in animal breeding, not the exceptional one.** A contemporary group
effect plus an overall mean is exactly the structure above. So is a sex effect with a mean, and a
breed effect with a mean. Chapter 2 is largely about what to do with it, and the answer — §1.10 —
is not to prevent it but to solve it anyway and be careful about which quantities you then quote.

### The determinant

The **determinant** is a single number computed from a square matrix, written $|\mathbf{A}|$ or
$\det(\mathbf{A})$. For a 2 × 2 it is the difference of the two diagonal products:

$$
\det \begin{bmatrix} a & b \\ c & d \end{bmatrix} = ad - bc
$$

For the full-rank $\mathbf{X}'\mathbf{X}$ of §1.5, $\det = (4)(2) - (2)(2) = 4$. For a matrix with
a redundant column the determinant is 0 — the 3 × 3 above has $\det = 0$, and so does the 2 × 2
$\begin{bmatrix} 100 & 100 \\ 100 & 100 \end{bmatrix}$, whose determinant is
$100(100) - 100(100) = 0$.

`det()` does all three:

```{r}
#| label: ch01-determinant
XtX <- crossprod(X)
XtX
(XtX[1, 1] * XtX[2, 2]) - (XtX[1, 2] * XtX[2, 1])   # ad - bc, by hand
det(XtX)                                            # the same, 4

det(XtX3)          # 0 -- the rank-deficient one
```

Do not memorise the general formula; you will never compute a determinant larger than 3 × 3 by
hand, and R does it in one call. Memorise the meaning instead, in its two forms:

- **For a design matrix**, $\det(\mathbf{X}'\mathbf{X}) = 0$ means a column is redundant and the
  effects are not separately estimable.
- **For a covariance matrix**, $\det(\mathbf{G}) = 0$ means a genetic correlation of exactly
  $\pm 1$ — one trait carries no information the other lacks. The matrix
  $\begin{bmatrix} 100 & 100 \\ 100 & 100 \end{bmatrix}$ says two traits with $\sigma^2_a = 100$
  each and $r_g = 1$: genetically, the same trait measured twice. REML will not converge on such a
  matrix, and Chapter 21 explains why.

Here is that covariance matrix — two traits with $\sigma^2_a = 100$ each and a genetic
correlation of exactly 1:

```{r}
#| label: ch01-singular-G
Gsing <- matrix(c(100, 100,
                  100, 100), nrow = 2)
Gsing

det(Gsing)          # 0
qr(Gsing)$rank      # 1, in a 2 x 2 matrix

# The genetic correlation this matrix implies:
Gsing[1, 2] / sqrt(Gsing[1, 1] * Gsing[2, 2])   # exactly 1
```

And this is what R does when you ask it to invert one. `solve()` computes an inverse — §1.9 is
where that is explained — and here it is deliberately being asked for one that does not exist:

```{r}
#| label: ch01-singular-fails
#| error: true
solve(Gsing)
```

**That message is the one to recognise.** "System is exactly singular" is R telling you a column
of your matrix is redundant. When it appears in a real analysis it almost always means two effects
in the model cannot be told apart from each other — not that anything is broken.

## Inversion, solving, and partitioned matrices

### Finding an inverse without a formula

The **inverse** of a square matrix $\mathbf{A}$ is the matrix $\mathbf{A}^{-1}$ for which
$\mathbf{AA}^{-1} = \mathbf{A}^{-1}\mathbf{A} = \mathbf{I}$. That definition is not just a
description — it is enough to *find* one.

Take $\mathbf{X}'\mathbf{X} = \begin{bmatrix} 4 & 2 \\ 2 & 2 \end{bmatrix}$ from §1.5, call the
unknown inverse $\begin{bmatrix} w & x \\ z & v \end{bmatrix}$, and demand that the product be the
identity:

$$
\begin{bmatrix} 4 & 2 \\ 2 & 2 \end{bmatrix}
\begin{bmatrix} w & x \\ z & v \end{bmatrix}
=
\begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix}
$$

Multiplying out, one element of the identity at a time, gives four equations in four unknowns:

$$
\begin{aligned}
4w + 2z &= 1 \qquad & 4x + 2v &= 0 \\
2w + 2z &= 0 \qquad & 2x + 2v &= 1
\end{aligned}
$$

The left pair: subtract the second from the first to get $2w = 1$, so $w = 0.5$, and then
$z = -0.5$. The right pair: subtract to get $2x = -1$, so $x = -0.5$, and then $v = 1$.

$$
(\mathbf{X}'\mathbf{X})^{-1} = \begin{bmatrix} 0.5 & -0.5 \\ -0.5 & 1 \end{bmatrix}
$$

Multiply back to confirm:
$\begin{bmatrix} 4 & 2 \\ 2 & 2 \end{bmatrix}\begin{bmatrix} 0.5 & -0.5 \\ -0.5 & 1 \end{bmatrix}
= \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix}$. It is the inverse.

<!-- {{< animation ch01-inverse-2x2 caption="AA-inverse = I, posed as four equations in four unknowns." >}} -->

**What mattered and what did not.** The particular entries 4, 2, 2, 2 set the answers, obviously.
But look at the shape of what came out. Every entry is something over 4 — and 4 is
$\det(\mathbf{X}'\mathbf{X})$, computed in §1.8. The two diagonal entries traded places (the 4 and
the 2 in the original became a 2/4 and a 4/4). The two off-diagonal entries kept their positions
but changed sign. None of that depended on the numbers; solve those four equations with any $a$,
$b$, $c$, $d$ and the same three things happen.

::: {.callout-important .key-equation}
## Key Equation — The inverse of a 2 × 2

$$
\begin{bmatrix} a & b \\ c & d \end{bmatrix}^{-1}
= \frac{1}{ad - bc}\begin{bmatrix} d & -b \\ -c & a \end{bmatrix}
$$ {#eq-inverse2}

**In words:** swap the diagonal entries, flip the sign of the off-diagonal ones, and divide
everything by the determinant. In general
$\mathbf{A}^{-1} = \operatorname{adj}(\mathbf{A})/\det(\mathbf{A})$, of which this is the 2 × 2
case — the only one you will ever compute by hand.
:::

The division makes §1.8 obvious in hindsight. If $\det(\mathbf{A}) = 0$ there is nothing to divide
by, and the inverse does not exist. "Singular" and "not invertible" are the same statement.

::: {.callout-note .definition}
## Definition — Inverse

The **inverse** of a square matrix $\mathbf{A}$, written $\mathbf{A}^{-1}$, is the matrix for
which $\mathbf{AA}^{-1} = \mathbf{A}^{-1}\mathbf{A} = \mathbf{I}$. It is the matrix equivalent
of a reciprocal, and it exists only when $\mathbf{A}$ is non-singular.

**Why it matters.** Inverting is how a matrix equation gets solved, and $\mathbf{A}^{-1}$ — the
inverse of the relationship matrix — is the single object that makes the animal model practical
(Chapter 4). Note the two cautions that follow: do not compute an inverse merely to solve a system
(below), and remember that the inverse of a covariance matrix is a *precision* matrix, which is
why relationships enter Henderson's equations as information rather than as covariance.
:::

### Three inverses worth recognising

The inverse of a **diagonal** matrix is the diagonal matrix of reciprocals — nothing else to do:

$$
\begin{bmatrix} 4 & 0 \\ 0 & 10 \end{bmatrix}^{-1}
= \begin{bmatrix} 0.25 & 0 \\ 0 & 0.1 \end{bmatrix}
$$

The inverse of a **product** also has a rule, and you can predict it. Look at @eq-reversal, then
write down what you think $(\mathbf{AB})^{-1}$ equals before reading the next sentence. — It is
$\mathbf{B}^{-1}\mathbf{A}^{-1}$, and for the same reason: the ends have to keep meeting.

The inverse of a **covariance** matrix is a **precision** matrix, and it is worth pausing on. A
covariance matrix says how much things vary together; its inverse says how much information you
have about each after accounting for the others. This is the first hint of something that looks
strange in Chapter 6: it is $\mathbf{A}^{-1}$, not $\mathbf{A}$, that appears in the mixed model
equations — and Henderson's rules for writing it down straight from a pedigree, without ever
forming $\mathbf{A}$ at all [@henderson1976simple], are what made the animal model computable.
Relationships enter the equations as information, not as covariance.

### Do not invert to solve

We want $\mathbf{b}$ from $\mathbf{X}'\mathbf{X}\,\mathbf{b} = \mathbf{X}'\mathbf{y}$. Having the
inverse, that is one multiplication:

$$
\hat{\mathbf{b}} = \begin{bmatrix} 0.5 & -0.5 \\ -0.5 & 1 \end{bmatrix}
\begin{bmatrix} 186 \\ 100 \end{bmatrix}
= \begin{bmatrix} 0.5(186) - 0.5(100) \\ -0.5(186) + 1(100) \end{bmatrix}
= \begin{bmatrix} 43 \\ 7 \end{bmatrix}
$$

Flock B's mean is 43 kg and flock A runs 7 kg above it — which matches the raw means, 50 and 43,
as it must for a design this balanced.

**But do not compute an inverse in order to solve a system.** `solve(A, b)` factorises and
back-substitutes; `solve(A) %*% b` forms the entire inverse first and then multiplies. The two
agree to rounding, but the second is slower, uses more memory, and is less numerically accurate,
and the gap grows with size. Chapter 24 makes this quantitative. Form an inverse only when you
want the inverse itself — which does happen, because the diagonal of
$(\mathbf{X}'\mathbf{X})^{-1}$ carries the standard errors, and in Chapter 6 the diagonal of the
inverted coefficient matrix carries prediction error variance and accuracy. When you do need it for
a symmetric matrix, `chol2inv(chol(A))` is the cheaper route: it inverts from the Cholesky factor
of §1.12 rather than starting over.

Both give 43 and 7 here. The difference shows up at scale, so time them on a 500 x 500 system:

```{r}
#| label: ch01-timing
set.seed(2024)
big_X <- matrix(rnorm(600 * 500), nrow = 600, ncol = 500)
big_A <- crossprod(big_X)          # 500 x 500, symmetric, full rank
big_b <- rnorm(500)

time_solve  <- system.time(solve(big_A, big_b))
time_invert <- system.time(solve(big_A) %*% big_b)

time_solve[["elapsed"]]     # solve(A, b)
time_invert[["elapsed"]]    # solve(A) %*% b
```

### Partitioned matrices

The mixed model equations are a 2 × 2 system of *blocks*, not of scalars, so it is worth seeing
that the 2 × 2 inverse formula works one level up. Partition the very matrix we just inverted,
taking each block to be 1 × 1:

$$
\mathbf{X}'\mathbf{X} = \left[\begin{array}{c|c}
4 & 2 \\ \hline 2 & 2
\end{array}\right]
= \begin{bmatrix} \mathbf{C}_{11} & \mathbf{C}_{12} \\ \mathbf{C}_{21} & \mathbf{C}_{22}
\end{bmatrix}
$$

The key quantity is the **Schur complement** of $\mathbf{C}_{11}$ — what is left of
$\mathbf{C}_{22}$ once $\mathbf{C}_{11}$'s contribution has been taken out of it:

$$
\mathbf{S} = \mathbf{C}_{22} - \mathbf{C}_{21}\mathbf{C}_{11}^{-1}\mathbf{C}_{12}
= 2 - 2\left(\tfrac{1}{4}\right)2 = 1
$$

And $\mathbf{S}^{-1} = 1$ is exactly the bottom-right entry of the inverse we found by hand. The
other blocks follow: $-\mathbf{C}_{11}^{-1}\mathbf{C}_{12}\mathbf{S}^{-1} =
-\tfrac{1}{4}(2)(1) = -0.5$ off the diagonal, and
$\mathbf{C}_{11}^{-1} + \mathbf{C}_{11}^{-1}\mathbf{C}_{12}\mathbf{S}^{-1}\mathbf{C}_{21}
\mathbf{C}_{11}^{-1} = 0.25 + 0.25 = 0.5$ top-left. The same inverse, reached a different way.

Nothing in that argument used the fact that the blocks were 1 × 1. When Chapter 7 absorbs the
animal equations into the fixed-effect equations, and when Chapter 2 absorbs contemporary groups,
they are computing a Schur complement.

::: {.callout-note .derivation collapse="true"}
## Derivation — the inverse of a partitioned matrix (optional)

Skip this on a first read. You already have the result above, verified on numbers; this shows it
holds for blocks of any size.

Write $\mathbf{C} = \begin{bmatrix} \mathbf{C}_{11} & \mathbf{C}_{12} \\
\mathbf{C}_{21} & \mathbf{C}_{22} \end{bmatrix}$ with $\mathbf{C}_{11}$ square and invertible, and
factor it into a lower-triangular, a block-diagonal, and an upper-triangular piece:

$$
\mathbf{C} =
\begin{bmatrix} \mathbf{I} & \mathbf{0} \\
\mathbf{C}_{21}\mathbf{C}_{11}^{-1} & \mathbf{I} \end{bmatrix}
\begin{bmatrix} \mathbf{C}_{11} & \mathbf{0} \\ \mathbf{0} & \mathbf{S} \end{bmatrix}
\begin{bmatrix} \mathbf{I} & \mathbf{C}_{11}^{-1}\mathbf{C}_{12} \\
\mathbf{0} & \mathbf{I} \end{bmatrix},
\qquad
\mathbf{S} = \mathbf{C}_{22} - \mathbf{C}_{21}\mathbf{C}_{11}^{-1}\mathbf{C}_{12}
$$

Multiply the three out to confirm the factorisation. Now invert, using
$(\mathbf{ABC})^{-1} = \mathbf{C}^{-1}\mathbf{B}^{-1}\mathbf{A}^{-1}$ from §1.9 — the reversal rule
again — and noting that a unit-triangular block matrix inverts by flipping the sign of its one
off-diagonal block:

$$
\mathbf{C}^{-1} =
\begin{bmatrix}
\mathbf{C}_{11}^{-1} + \mathbf{C}_{11}^{-1}\mathbf{C}_{12}\mathbf{S}^{-1}\mathbf{C}_{21}\mathbf{C}_{11}^{-1}
& -\mathbf{C}_{11}^{-1}\mathbf{C}_{12}\mathbf{S}^{-1} \\
-\mathbf{S}^{-1}\mathbf{C}_{21}\mathbf{C}_{11}^{-1} & \mathbf{S}^{-1}
\end{bmatrix}
$$

Setting every block to a scalar recovers @eq-inverse2. Two consequences used later: the
bottom-right block of the inverse depends on $\mathbf{C}_{22}$ only through the Schur complement,
which is why prediction error variance for the animals accounts for the fixed effects having been
estimated from the same data; and $\det(\mathbf{C}) = \det(\mathbf{C}_{11})\det(\mathbf{S})$,
which is how Chapter 21 evaluates a REML likelihood without forming the whole determinant.
:::

```{r}
#| label: ch01-inverse-solve
XtX <- crossprod(X)
Xty <- crossprod(X, y)

solve(XtX)              # the inverse we found by hand
solve(XtX, Xty)         # b-hat = 43, 7 -- solve directly, the right way
solve(XtX) %*% Xty      # same answer, by forming the inverse first
```

## Generalized inverses

The rank-deficient $\mathbf{X}_3'\mathbf{X}_3$ of §1.8 has no inverse, but it still has
solutions — infinitely many of them, and a way of reaching one.

::: {.callout-note .definition}
## Definition — Generalized inverse

A **generalized inverse** of $\mathbf{A}$, written $\mathbf{A}^-$, is any matrix satisfying
$\mathbf{AA}^-\mathbf{A} = \mathbf{A}$. That single condition is enough to produce a solution to
the system, and it is much weaker than the condition defining a true inverse — which is why a
singular matrix has generalized inverses even though it has no inverse.

Crucially, $\mathbf{A}^-$ is **not unique**. A singular matrix has infinitely many, and different
ones give different solution vectors.

**Why it matters.** Every fixed-effect model in this book is solved with a generalized inverse,
because every one of them is rank deficient. The consequence is the one thing to carry out of this
section: the individual numbers in $\hat{\mathbf{b}}$ depend on which generalized inverse you
happened to use, so they are not results. What you can report is defined next.
:::

The easiest way to build one is to drop enough columns to leave a full-rank piece, invert that,
and pad the result back out with zeros. There are three obvious choices here; take two of them.

**Drop the baseline.** Keep only the two flock columns, invert their 2 × 2 corner
$\begin{bmatrix} 2 & 0 \\ 0 & 2 \end{bmatrix}$, and pad. The solution is

$$
\hat{\mathbf{b}}_1 = (0,\ 50,\ 43)'
$$

Built in R, the padding is what makes it a *generalized* inverse — the baseline's row and column
stay zero:

```{r}
#| label: ch01-ginv-build
XtX3 <- crossprod(X3)
Xty3 <- crossprod(X3, y)

Ginv_drop_baseline <- matrix(0, nrow = 3, ncol = 3)
Ginv_drop_baseline[2:3, 2:3] <- solve(XtX3[2:3, 2:3])
Ginv_drop_baseline

drop(Ginv_drop_baseline %*% Xty3)     # 0, 50, 43
```

**Drop the flock B column.** Keep the baseline and flock A, which is exactly the $\mathbf{X}$ of
§1.7, and pad the zero back in at the end:

$$
\hat{\mathbf{b}}_2 = (43,\ 7,\ 0)'
$$

```{r}
#| label: ch01-ginv-build-2
Ginv_drop_flockB <- matrix(0, nrow = 3, ncol = 3)
Ginv_drop_flockB[1:2, 1:2] <- solve(XtX3[1:2, 1:2])

drop(Ginv_drop_flockB %*% Xty3)       # 43, 7, 0
```

Both really are generalized inverses. The defining property is
$\mathbf{AA}^-\mathbf{A} = \mathbf{A}$, and it holds for each:

```{r}
#| label: ch01-ginv-property
all.equal(XtX3 %*% Ginv_drop_baseline %*% XtX3, XtX3)
all.equal(XtX3 %*% Ginv_drop_flockB   %*% XtX3, XtX3)
```

These are not small disagreements. In the first, flock A's effect is 50 and the baseline is 0; in
the second, flock A's effect is 7 and the baseline is 43. **Neither is more correct than the
other**, and asking "what is the effect of flock A?" has no answer.

Now compute the fitted values from each. Both give $(50, 50, 43, 43)'$ — the two flock means,
identically. And both give the same flock difference:
$50 - 43 = 7$ from the first, $7 - 0 = 7$ from the second.

```{r}
#| label: ch01-ginv-fitted
b1 <- Ginv_drop_baseline %*% Xty3
b2 <- Ginv_drop_flockB   %*% Xty3

# Different solutions...
rbind(drop_baseline = drop(b1), drop_flockB = drop(b2))

# ...identical fitted values.
rbind(drop_baseline = drop(X3 %*% b1), drop_flockB = drop(X3 %*% b2))

# ...and an identical flock difference.
c(from_first = drop(b1)[2] - drop(b1)[3],
  from_second = drop(b2)[2] - drop(b2)[3])
```

::: {.callout-caution .check collapse="true"}
## Check Yourself

Two different solution vectors, same fitted values. Which quantities can you report, and which
can you not?

**Answer.** You can report anything that came out the same: the fitted values, the flock
difference of 7 kg, the residuals, and any error variance computed from them. You cannot report
the individual elements of $\hat{\mathbf{b}}$, because they changed. The general test is that a
quantity is trustworthy when it is a **linear function of the data's expected values** — a
combination $\mathbf{k}'\mathbf{b}$ for which $\mathbf{k}'$ can be written as some combination of
rows of $\mathbf{X}$. Such a combination is called **estimable**, and Chapter 2 makes it precise.
Here $(0, 1, -1)$ is estimable and $(0, 1, 0)$ is not.
:::

::: {.callout-note .definition}
## Definition — Estimable function

A linear combination $\mathbf{k}'\mathbf{b}$ of the effects is **estimable** when its value is
the same no matter which generalized inverse was used to obtain $\hat{\mathbf{b}}$. The formal
test is that $\mathbf{k}'$ must be expressible as a linear combination of the rows of
$\mathbf{X}$.

Here the flock difference $(0, 1, -1)$ is estimable and comes to 7 kg every time; the individual
flock A effect $(0, 1, 0)$ is not estimable and takes a different value under every solution.

**Why it matters.** Estimability is the line between a number you may publish and a number that is
an artefact of your software's internal choices. Differences between levels, fitted values, and
residuals are estimable; individual effects in a rank-deficient model are not. Chapter 2 makes the
test precise, and Chapter 9 returns to it when genetic groups make the question harder.
:::

| Solution | $\hat{b}_\mu$ | $\hat{b}_A$ | $\hat{b}_B$ | Fitted values | $\hat{b}_A - \hat{b}_B$ |
|---|---|---|---|---|---|
| Drop the baseline | 0 | 50 | 43 | 50, 50, 43, 43 | 7 |
| Drop flock B | 43 | 7 | 0 | 50, 50, 43, 43 | 7 |
| `MASS::ginv` (Moore–Penrose) | 31 | 19 | 12 | 50, 50, 43, 43 | 7 |

: Three generalized inverses of the same $\mathbf{X}_3'\mathbf{X}_3$. The individual solutions
disagree completely; everything a breeder would report agrees exactly. {#tbl-ginv}

@tbl-ginv collects all three. The third row is what R's `MASS::ginv()` produces — the Moore–Penrose inverse [@penrose1955generalized], which adds three
further conditions to $\mathbf{AA}^-\mathbf{A} = \mathbf{A}$ and so is unique. It is a perfectly
good g-inverse and it gives a third completely different answer, which is the point of the table.
What `lm()` does is the second row: it silently drops a level and reports the rest relative to it.

```{r}
#| label: ch01-ginv-moore-penrose
# MASS ships with R, so nothing needs installing.
b3 <- MASS::ginv(XtX3) %*% Xty3

rbind(drop_baseline = drop(b1),
      drop_flockB   = drop(b2),
      moore_penrose = drop(b3))

# Still the same fitted values, and still a 7 kg difference.
drop(X3 %*% b3)
drop(b3)[2] - drop(b3)[3]
```

R's own `lm()` takes the second route — it silently drops a level and reports everything relative
to it. That is why an `lm()` summary shows one fewer coefficient than you have levels:

```{r}
#| label: ch01-lm-drops-a-level
coef(lm(wt ~ flock, data = lambs))
```

## Kronecker product, Hadamard product, and direct sum

Three products that are not matrix multiplication. Each builds a big matrix out of small ones, and
each has exactly one job in this book.

### Kronecker product ⊗ — multiple traits

Two full sibs, two traits. The relationship matrix for the sibs and the genetic covariance matrix
for the traits are both 2 × 2:

$$
\mathbf{A} = \begin{bmatrix} 1 & 0.5 \\ 0.5 & 1 \end{bmatrix}
\qquad
\mathbf{G}_0 = \begin{bmatrix} 36 & 12 \\ 12 & 9 \end{bmatrix}
$$

Two animals with two traits each means four breeding values, so their covariance matrix must be
4 × 4. It is built by replacing each element of one matrix with that element times the whole of
the other. Taking $\mathbf{G}_0$ as the outer matrix:

$$
\mathbf{G}_0 \otimes \mathbf{A} =
\left[\begin{array}{cc|cc}
36 & 18 & 12 & 6 \\
18 & 36 & 6 & 12 \\ \hline
12 & 6 & 9 & 4.5 \\
6 & 12 & 4.5 & 9
\end{array}\right]
$$

<!-- {{< animation ch01-kronecker caption="G0 (x) A expands a 2x2 and a 2x2 into a 4x4." >}} -->

In R the operator is `%x%`, and `kronecker()` is the same function spelled out:

```{r}
#| label: ch01-kronecker-build
# From here A is the relationship matrix, not the generic 2 x 2 of section 1.3.
A  <- matrix(c(1, 0.5,
               0.5, 1), nrow = 2)     # two full sibs
G0 <- matrix(c(36, 12,
               12,  9), nrow = 2)     # the two traits

G0 %x% A                              # 4 x 4: animal within trait
dim(G0 %x% A)
```

Read it. The top-left block is $36\mathbf{A}$ — trait 1's genetic variance spread over the two
sibs. The bottom-right is $9\mathbf{A}$, trait 2's. The off-diagonal blocks are
$12\mathbf{A}$, holding every trait-1-with-trait-2 covariance: 12 on the block's diagonal, which is
one animal's two traits, and $12 \times 0.5 = 6$ off it, which is one animal's trait 1 with the
*other* animal's trait 2 — halved, because sibs share half their additive genes.

**The ordering matters and mixing the two up is the classic multi-trait error.** Written as
$\mathbf{G}_0 \otimes \mathbf{A}$, the rows run trait 1 for both animals, then trait 2 for both —
**animal within trait**. Written the other way round it is **trait within animal**:

$$
\mathbf{A} \otimes \mathbf{G}_0 =
\left[\begin{array}{cc|cc}
36 & 12 & 18 & 6 \\
12 & 9 & 6 & 4.5 \\ \hline
18 & 6 & 36 & 12 \\
6 & 4.5 & 12 & 9
\end{array}\right]
$$

```{r}
#| label: ch01-kronecker-ordering
A %x% G0        # the other ordering: trait within animal
```

Compare element (1, 2) between the two: 18 in the first, 12 in the second. Same sixteen numbers,
different positions — which is exactly the error the box below warns about.

The general definition is that
$\mathbf{A} \otimes \mathbf{B}$ replaces each $a_{ij}$ with the block $a_{ij}\mathbf{B}$, so an
$m \times n$ times a $p \times q$ gives $mp \times nq$. One property is worth carrying forward:

$$
(\mathbf{A} \otimes \mathbf{B})(\mathbf{C} \otimes \mathbf{D}) = \mathbf{AC} \otimes \mathbf{BD}
$$

The consequence worth checking yourself: you can invert the two small factors instead of the big
product, which is the difference between inverting two 2 × 2s and one 4 × 4 — or, in Chapter 11,
between a handful of traits and every animal in the pedigree.

```{r}
#| label: ch01-kronecker-inverse
all.equal(solve(G0 %x% A), solve(G0) %x% solve(A))
```

That mixed-product rule is what keeps multi-trait mixed model equations tractable — it lets you
invert a Kronecker product by inverting its two small factors separately, since
$(\mathbf{G}_0 \otimes \mathbf{A})^{-1} = \mathbf{G}_0^{-1} \otimes \mathbf{A}^{-1}$. Chapter 11
lives on this.

::: {.callout-warning}
## Common Mistake — trait within animal versus animal within trait

Both orderings are correct; they simply describe different arrangements of the same solution
vector. The error is using one ordering for $\mathbf{G}_0 \otimes \mathbf{A}$ and the other for
$\mathbf{Z}$ or for the solution vector, which silently scrambles the covariances rather than
raising an error.

Fix the ordering once, write it down beside the model, and check it by picking one off-diagonal
element and saying out loud which two breeding values it is the covariance of. In
$\mathbf{G}_0 \otimes \mathbf{A}$ above, element (1, 4) is 6 — the covariance between animal 1's
trait 1 and animal 2's trait 2. In $\mathbf{A} \otimes \mathbf{G}_0$ that same covariance sits at
element (1, 4) as well, but for a different reason, and element (1, 2) has changed from 18 to 12.
:::

### Hadamard product # — non-additive effects

The Hadamard product multiplies element by element, so both matrices must be the same size and the
result is that size too. For the two full sibs:

$$
\mathbf{A} \# \mathbf{A} =
\begin{bmatrix} 1 & 0.5 \\ 0.5 & 1 \end{bmatrix} \#
\begin{bmatrix} 1 & 0.5 \\ 0.5 & 1 \end{bmatrix} =
\begin{bmatrix} 1 & 0.25 \\ 0.25 & 1 \end{bmatrix}
$$

```{r}
#| label: ch01-hadamard
A * A        # element by element -- NOT matrix multiplication
```

Full sibs share half their additive genes but only a quarter of their additive-by-additive
combinations, and the squaring is where that quarter comes from. $\mathbf{A} \# \mathbf{A}$ is the
relationship matrix for additive × additive epistasis and $\mathbf{A} \# \mathbf{D}$ for
additive × dominance; Chapter 17 uses both, and Chapter 16 builds the dominance matrix
$\mathbf{D}$ they need.

### Direct sum ⊕ — independent blocks

The direct sum stacks matrices into a block-diagonal arrangement with zeros everywhere else:

$$
\mathbf{A}_1 \oplus \mathbf{A}_2 =
\begin{bmatrix} \mathbf{A}_1 & \mathbf{0} \\ \mathbf{0} & \mathbf{A}_2 \end{bmatrix}
$$

`Matrix::bdiag()` builds one. Two unrelated sib pairs have no covariance across the blocks, so the
off-diagonal corners are zero:

```{r}
#| label: ch01-direct-sum
as.matrix(bdiag(A, A))
```

This is what a set of unrelated families looks like, and what a residual covariance matrix looks
like when animals are grouped into contemporary groups with no covariance across groups
(Chapters 5 and 24).

## Cholesky decomposition

Take the covariance matrix

$$
\mathbf{M} = \begin{bmatrix} 100 & 60 \\ 60 & 100 \end{bmatrix}
$$

Call this **Example C**. It is used again in §1.13.

and ask for a lower-triangular $\mathbf{L}$ with $\mathbf{M} = \mathbf{LL}'$. Write the unknown
$\mathbf{L}$ out and multiply:

$$
\begin{bmatrix} \ell_{11} & 0 \\ \ell_{21} & \ell_{22} \end{bmatrix}
\begin{bmatrix} \ell_{11} & \ell_{21} \\ 0 & \ell_{22} \end{bmatrix}
=
\begin{bmatrix} \ell_{11}^2 & \ell_{11}\ell_{21} \\
\ell_{11}\ell_{21} & \ell_{21}^2 + \ell_{22}^2 \end{bmatrix}
$$

Now match entries against $\mathbf{M}$, in order, and watch what each step needs:

1. Top-left: $\ell_{11}^2 = 100$, so $\ell_{11} = 10$.
2. Below it: $\ell_{11}\ell_{21} = 60$. We already have $\ell_{11} = 10$, so $\ell_{21} = 6$.
3. Bottom-right: $\ell_{21}^2 + \ell_{22}^2 = 100$. We already have $\ell_{21} = 6$, so
   $\ell_{22}^2 = 100 - 36 = 64$ and $\ell_{22} = 8$.

$$
\mathbf{L} = \begin{bmatrix} 10 & 0 \\ 6 & 8 \end{bmatrix}
$$

<!-- {{< animation ch01-cholesky caption="M = LL', one entry at a time, each using the last." >}} -->

**What mattered and what did not.** The numbers set the answers, but the *dependency* is the whole
idea and it did not depend on the numbers at all: step 2 could not start until step 1 finished, and
step 3 needed step 2. That is why this is called a recursion rather than a formula. Written out for
a matrix of any size, each entry uses only entries already computed:

$$
\ell_{jj} = \sqrt{m_{jj} - \sum_{k=1}^{j-1}\ell_{jk}^2}
\qquad
\ell_{ij} = \frac{1}{\ell_{jj}}\left(m_{ij} - \sum_{k=1}^{j-1}\ell_{ik}\ell_{jk}\right), \quad i > j
$$

Note the square root. If any diagonal entry comes out as the square root of a negative number, the
matrix is not positive definite and the factorisation fails — which makes Cholesky the standard
*test* for whether a proposed covariance matrix is valid at all. That connects directly to §1.13.

Three uses, each named with its chapter:

1. **Solving.** Factor once, then solve two triangular systems by substitution — `chol()`
   followed by `backsolve()`. For a symmetric positive definite system such as
   $\mathbf{X}'\mathbf{X}\,\mathbf{b} = \mathbf{X}'\mathbf{y}$ this is the numerically sound
   route, and it is roughly twice as fast as the general-purpose LU factorisation that plain
   `solve()` uses (Chapter 24).
2. **Simulating correlated breeding values.** If $\mathbf{z}$ is a vector of independent standard
   normals, then $\mathbf{a} = \mathbf{Lz}$ has covariance matrix $\mathbf{M}$ — because
   $\mathbf{L}\,\mathbf{I}\,\mathbf{L}' = \mathbf{LL}' = \mathbf{M}$. Section 1.15 works out
   why that is the right calculation; for now, the R chunk below simply checks it on 10,000
   samples. Every simulated two-trait animal in this book is generated this way (Chapters 11
   and 22).
3. **Log-determinants.** $\log|\mathbf{M}| = 2\sum_j \log \ell_{jj}$, which is how a REML
   likelihood gets evaluated without ever forming a determinant directly (Chapter 21).

A fourth connection, for later: Chapter 4 factors the relationship matrix as
$\mathbf{A} = \mathbf{TDT}'$, which is a Cholesky-like factorisation whose triangular factor has a
direct genetic reading — row $i$ of $\mathbf{T}$ says what fraction of animal $i$'s genes came from
each ancestor. That is where $\mathbf{A}^{-1}$'s famous sparsity comes from.

::: {.callout-tip}
## In R — `chol()` returns the **upper** triangle

R's `chol()` returns the upper-triangular $\mathbf{R}$ with $\mathbf{M} = \mathbf{R}'\mathbf{R}$,
not the lower-triangular $\mathbf{L}$ with $\mathbf{M} = \mathbf{LL}'$ that the textbooks write.
They are transposes of each other, so `L <- t(chol(M))`.

Getting this wrong is a genuine trap because `chol(M) %*% t(chol(M))` runs without error and
returns a symmetric matrix that is simply not $\mathbf{M}$. Simulating with `chol(G) %*% z` instead
of `t(chol(G)) %*% z` gives breeding values with the wrong covariance structure and no warning at
all. Check with `all.equal(L %*% t(L), M)` every time.
:::

```{r}
#| label: ch01-cholesky
M <- matrix(c(100, 60,
               60, 100), nrow = 2)

chol(M)           # R, UPPER triangular: 10 6 / 0 8
L <- t(chol(M))   # transpose it to get the L of the hand calculation
L

# Always check: LL' must give M back.
all.equal(L %*% t(L), M)
```

Now use $\mathbf{L}$ to simulate. Draw independent standard normals, multiply by
$\mathbf{L}$, and the results carry $\mathbf{M}$'s covariance structure:

```{r}
#| label: ch01-cholesky-sim
set.seed(2024)
z <- matrix(rnorm(2 * 10000), nrow = 2)   # 10,000 pairs, independent
a <- L %*% z                              # now correlated

round(cov(t(a)), 1)    # recovers M: 100, 60 / 60, 100
```

And the log-determinant, twice over — from the diagonal of $\mathbf{L}$, and directly:

```{r}
#| label: ch01-logdet
2 * sum(log(diag(L)))
log(det(M))
```

## Eigenvalues and eigenvectors

An **eigenvector** of a square matrix is a direction that the matrix does not rotate — it only
stretches it. The stretch factor is the **eigenvalue**: $\mathbf{Mv} = \lambda\mathbf{v}$.

For Example C's $\mathbf{M}$, try the direction $(1, 1)'$, meaning "both traits together":

$$
\begin{bmatrix} 100 & 60 \\ 60 & 100 \end{bmatrix}
\begin{bmatrix} 1 \\ 1 \end{bmatrix}
= \begin{bmatrix} 160 \\ 160 \end{bmatrix}
= 160 \begin{bmatrix} 1 \\ 1 \end{bmatrix}
$$

Same direction, stretched by 160. Now try $(1, -1)'$, meaning "one trait up, the other down":

$$
\begin{bmatrix} 100 & 60 \\ 60 & 100 \end{bmatrix}
\begin{bmatrix} 1 \\ -1 \end{bmatrix}
= \begin{bmatrix} 40 \\ -40 \end{bmatrix}
= 40 \begin{bmatrix} 1 \\ -1 \end{bmatrix}
$$

Same direction, stretched by 40. Those are the two eigenvalues, 160 and 40, and their eigenvectors
are conventionally scaled to unit length as $(1,1)'/\sqrt{2}$ and $(1,-1)'/\sqrt{2}$.

The interpretation is the reason to care. The two traits are positively correlated, so most of the
variation is animals that are high for both or low for both — that is the 160 axis. Much less
variation is animals high for one and low for the other — the 40 axis. Drawn as an ellipse, the
eigenvectors are the axes and the square roots of the eigenvalues are their half-lengths.

```{r}
#| label: fig-ellipse
#| fig-cap: "The covariance ellipse of Example C, with its eigenvectors drawn as axes. The long axis is the 'both traits together' direction (eigenvalue 160); the short axis is the contrast direction (eigenvalue 40)."
#| fig-width: 5.5
#| fig-height: 5.5
#| out-width: "68%"
#| code-fold: true
# Folded: this is plot cosmetics, not algebra. The eigen-decomposition itself
# is in the visible chunk below.
eig <- eigen(M)
angle <- seq(0, 2 * pi, length.out = 400)
pts <- eig$vectors %*% diag(sqrt(eig$values)) %*% rbind(cos(angle), sin(angle))

par(mar = c(4, 4, 1, 1))
plot(pts[1, ], pts[2, ], type = "l", asp = 1, lwd = 2,
     xlab = "trait 1 deviation", ylab = "trait 2 deviation")
abline(h = 0, v = 0, col = "grey80")
for (j in 1:2) {
  axis_j <- eig$vectors[, j] * sqrt(eig$values[j])
  arrows(0, 0, axis_j[1], axis_j[2], lwd = 2, length = 0.1,
         col = c("firebrick", "steelblue")[j])
  text(axis_j[1] * 1.18, axis_j[2] * 1.18,
       bquote(lambda == .(eig$values[j])), cex = 0.9)
}
```

::: {.callout-caution .check collapse="true"}
## Check Yourself

The eigenvalues of $\mathbf{M}$ are 160 and 40. Before computing it: what is
$\det(\mathbf{M})$, and what would the determinant be if one eigenvalue were 0?

**Answer.** $160 \times 40 = 6400$, and directly, $100(100) - 60(60) = 10000 - 3600 = 6400$. The
determinant is the product of the eigenvalues, so a single zero eigenvalue drives the whole
determinant to zero — which is §1.8's singularity seen from a different angle. Likewise the trace
(the sum of the diagonal, $100 + 100 = 200$) equals the sum of the eigenvalues,
$160 + 40 = 200$.
:::

Three facts a breeder needs:

- **All eigenvalues > 0 means positive definite**, which is the condition for a matrix to be a
  valid covariance matrix. It is the same condition Cholesky tests in §1.12.

::: {.callout-note .definition}
## Definition — Positive definite

A symmetric matrix $\mathbf{M}$ is **positive definite** when every one of its eigenvalues is
greater than zero. Equivalently, $\mathbf{x}'\mathbf{Mx} > 0$ for every non-zero vector
$\mathbf{x}$, and its Cholesky factorisation (§1.12) succeeds. If eigenvalues are zero but none
are negative it is **positive semi-definite**.

**Why it matters.** A covariance matrix must be positive definite to be a covariance matrix at
all. Since $\mathbf{x}'\mathbf{Mx}$ is the variance of a linear combination (§1.15), a
non-positive-definite $\mathbf{G}$ would give some index of traits a *negative* variance, which is
impossible. This is what your software means by "G is not positive definite" — a message you will
meet in Chapter 21, and one that usually says the data cannot support as many traits as you asked
for rather than that the program is broken.
:::
- **A zero eigenvalue means a perfect genetic correlation.** Example C's singular sibling
  $\begin{bmatrix} 100 & 100 \\ 100 & 100 \end{bmatrix}$ has eigenvalues 200 and 0: all the
  variation is on the "together" axis and none on the contrast axis, because $r_g = 1$ and the two
  traits are genetically the same trait. REML cannot estimate such a matrix (Chapter 21).
- **Eigenvectors are the principal components of $\mathbf{G}$.** When a genetic covariance matrix
  for twenty traits has most of its variance in three eigenvalues, you can model three dimensions
  instead of twenty. That is what reduced-rank and factor-analytic random regression models do
  (Chapter 12).

::: {.callout-note .derivation collapse="true"}
## Derivation — why symmetric matrices behave so well (optional)

Skip this on a first read. It explains why every eigenvalue in this book is a real number and
every pair of eigenvectors is at right angles — neither of which is true for matrices in general.

**Real eigenvalues.** Let $\mathbf{Mv} = \lambda\mathbf{v}$ with $\mathbf{M}$ real and symmetric,
allowing $\lambda$ and $\mathbf{v}$ to be complex for the moment. Pre-multiply by the conjugate
transpose $\bar{\mathbf{v}}'$:

$$
\bar{\mathbf{v}}'\mathbf{M}\mathbf{v} = \lambda\,\bar{\mathbf{v}}'\mathbf{v}
$$

The left side is its own conjugate transpose, because $\mathbf{M}$ is real and symmetric, so it is
real. And $\bar{\mathbf{v}}'\mathbf{v} = \sum_i |v_i|^2$ is real and positive. A real number
divided by a positive real number is real, so $\lambda$ is real.

**Orthogonal eigenvectors.** Take two eigenpairs with $\lambda_1 \neq \lambda_2$. Then

$$
\lambda_1 \mathbf{v}_2'\mathbf{v}_1 = \mathbf{v}_2'\mathbf{Mv}_1
= (\mathbf{M}\mathbf{v}_2)'\mathbf{v}_1 = \lambda_2 \mathbf{v}_2'\mathbf{v}_1
$$

using $\mathbf{M}' = \mathbf{M}$ in the middle step. So
$(\lambda_1 - \lambda_2)\mathbf{v}_2'\mathbf{v}_1 = 0$, and since the eigenvalues differ,
$\mathbf{v}_2'\mathbf{v}_1 = 0$. The eigenvectors are perpendicular — which is why the ellipse
above has axes at right angles, and why the eigen-decomposition
$\mathbf{M} = \mathbf{V}\boldsymbol{\Lambda}\mathbf{V}'$ has $\mathbf{V}^{-1} = \mathbf{V}'$.

**Consequences used later.** $\det(\mathbf{M}) = \prod_j \lambda_j$ and
$\operatorname{tr}(\mathbf{M}) = \sum_j \lambda_j$ both follow from that decomposition, and so does
the statement that $\mathbf{M}$ is positive definite exactly when every $\lambda_j > 0$.
:::

```{r}
#| label: ch01-eigen
eig <- eigen(M)
eig$values      # 160 and 40
eig$vectors     # the sum axis and the contrast axis, scaled to length 1

# The determinant is the product of the eigenvalues.
det(M)
prod(eig$values)

# The trace is their sum.
sum(diag(M))
sum(eig$values)

# All eigenvalues positive means positive definite: a valid covariance matrix.
all(eig$values > 0)

# The rg = 1 matrix is not: one eigenvalue is exactly zero.
eigen(matrix(100, nrow = 2, ncol = 2))$values
```

A closely related tool, the **singular value decomposition**, factors any matrix — square or not —
as $\mathbf{U}\mathbf{D}\mathbf{V}'$, and for a symmetric positive definite matrix it coincides
with the eigen-decomposition. It is the workhorse behind numerically stable rank determination and
behind the low-rank approximations used in large-scale evaluation. This book does not develop it;
see the companion volume on computation.

## Quadratic forms, traces, and matrix derivatives

**This section exists to serve Chapter 21.** Variance component estimation by REML is an exercise
in differentiating a likelihood with respect to variance components, and the likelihood is built
from quadratic forms and traces. Everything here is a tool you will pick up again there. Chapter 2
also needs the last part of it, three pages from now.

### Quadratic forms

A **quadratic form** is $\mathbf{x}'\mathbf{Ax}$ — a row vector, a matrix, and the same vector
again as a column. The result is always a single number. The simplest one in this book is the
residual sum of squares. From §1.9, the residuals of Example A are

$$
\mathbf{e} = \mathbf{y} - \mathbf{X}\hat{\mathbf{b}}
= \begin{bmatrix} 48 \\ 52 \\ 41 \\ 45 \end{bmatrix} -
  \begin{bmatrix} 50 \\ 50 \\ 43 \\ 43 \end{bmatrix}
= \begin{bmatrix} -2 \\ 2 \\ -2 \\ 2 \end{bmatrix}
$$

and

$$
\mathbf{e}'\mathbf{e} = (-2)^2 + 2^2 + (-2)^2 + 2^2 = 16
$$

```{r}
#| label: ch01-residuals
b_hat <- drop(solve(crossprod(X), crossprod(X, y)))
e <- drop(y - X %*% b_hat)
e

drop(crossprod(e))     # e'e = 16, the residual sum of squares
```

which is $\mathbf{e}'\mathbf{Ie}$, a quadratic form with $\mathbf{A} = \mathbf{I}$. Dividing by the
residual degrees of freedom, $4 - 2 = 2$, gives $\hat{\sigma}^2_e = 8$ kg².

```{r}
#| label: ch01-sigma2e
df <- nrow(X) - qr(X)$rank    # 4 records - 2 estimated effects
df
drop(crossprod(e)) / df       # sigma-hat squared = 8
```

The same construction gives $\mathbf{y}'\mathbf{y} = 8714$ as the uncorrected total sum of squares and
$\hat{\mathbf{b}}'\mathbf{X}'\mathbf{y} = 43(186) + 7(100) = 8698$ as the part explained by the
model, and $8714 - 8698 = 16$ recovers the residual sum of squares — the sums-of-squares partition
Chapter 20 estimates variance components from.

```{r}
#| label: ch01-ss-partition
yty      <- drop(crossprod(y))               # total
ss_model <- drop(b_hat %*% crossprod(X, y))  # explained by the model
c(total = yty, model = ss_model, residual = yty - ss_model)
```

### Trace

The **trace** of a square matrix is the sum of its diagonal, $\operatorname{tr}(\mathbf{A}) =
\sum_i a_{ii}$. It equals the sum of the eigenvalues (§1.13), and it has one property that gets
used constantly in REML because it permits a reordering that nothing else does:

$$
\operatorname{tr}(\mathbf{AB}) = \operatorname{tr}(\mathbf{BA})
$$

There is no `trace()` in base R; you sum the diagonal yourself. Check the reordering property on
two matrices whose products are plainly different:

```{r}
#| label: ch01-trace
AB <- A %*% G0
BA <- G0 %*% A
AB                                                      # not the same matrix
BA
c(trace_AB = sum(diag(AB)), trace_BA = sum(diag(BA)))   # but the same trace
```

The result worth memorising concerns the **hat matrix**
$\mathbf{H} = \mathbf{X}(\mathbf{X}'\mathbf{X})^{-}\mathbf{X}'$, which turns observations into
fitted values: $\hat{\mathbf{y}} = \mathbf{Hy}$. Its trace equals the rank of $\mathbf{X}$. For
Example A, $\operatorname{tr}(\mathbf{H}) = 2$, and the residual degrees of freedom are
$n - \operatorname{tr}(\mathbf{H}) = 4 - 2 = 2$ — the same 2 used above. **That is where degrees of
freedom come from**: they count the dimensions the model consumed, and the trace of the projection
matrix counts them whether or not $\mathbf{X}$ was full rank.

```{r}
#| label: ch01-hat-matrix
H <- X %*% solve(crossprod(X)) %*% t(X)

drop(H %*% y)             # the fitted values: 50 50 43 43
sum(diag(H))              # trace = 2
qr(X)$rank                # the same 2
nrow(X) - sum(diag(H))    # residual degrees of freedom
```

### Matrix derivatives

Two derivatives, both of a scalar with respect to a vector. The result has the same shape as the
vector you differentiated with respect to.

$$
\frac{\partial (\mathbf{a}'\mathbf{x})}{\partial \mathbf{x}} = \mathbf{a}
\qquad
\frac{\partial (\mathbf{x}'\mathbf{Ax})}{\partial \mathbf{x}} = 2\mathbf{Ax}
\quad (\mathbf{A} \text{ symmetric})
$$

The second is the matrix version of $\mathrm{d}(ax^2)/\mathrm{d}x = 2ax$, which is a good way to
remember it. Apply them immediately. Least squares chooses $\mathbf{b}$ to minimise the residual
sum of squares, which expands to

$$
(\mathbf{y} - \mathbf{Xb})'(\mathbf{y} - \mathbf{Xb})
= \mathbf{y}'\mathbf{y} - 2\mathbf{b}'\mathbf{X}'\mathbf{y}
+ \mathbf{b}'\mathbf{X}'\mathbf{Xb}
$$

Differentiate with respect to $\mathbf{b}$, using the first rule on the middle term and the second
on the last (note $\mathbf{X}'\mathbf{X}$ is symmetric), and set the result to zero:

$$
-2\mathbf{X}'\mathbf{y} + 2\mathbf{X}'\mathbf{Xb} = \mathbf{0}
\qquad \Longrightarrow \qquad
\mathbf{X}'\mathbf{X}\,\hat{\mathbf{b}} = \mathbf{X}'\mathbf{y}
$$

Those are the **normal equations**, and we have been solving them since §1.9 without having derived
them. Chapter 2 opens with the result already in hand.

::: {.callout-note .derivation collapse="true"}
## Derivation — the two matrix derivatives (optional)

Skip this on a first read. Both rules are just ordinary calculus done one element at a time.

**Linear form.** $\mathbf{a}'\mathbf{x} = \sum_k a_k x_k$. Differentiating with respect to a single
element $x_i$ kills every term except the one containing it:

$$
\frac{\partial}{\partial x_i}\sum_k a_k x_k = a_i
$$

Stacking those partials into a vector gives $\mathbf{a}$.

**Quadratic form.** $\mathbf{x}'\mathbf{Ax} = \sum_j \sum_k x_j a_{jk} x_k$. Element $x_i$ appears
in the terms where $j = i$ and in the terms where $k = i$, so the product rule gives

$$
\frac{\partial}{\partial x_i}\sum_j\sum_k x_j a_{jk} x_k
= \sum_k a_{ik}x_k + \sum_j x_j a_{ji}
$$

The first sum is row $i$ of $\mathbf{Ax}$; the second is row $i$ of $\mathbf{A}'\mathbf{x}$. In
general the derivative is therefore $(\mathbf{A} + \mathbf{A}')\mathbf{x}$, and when $\mathbf{A}$
is symmetric the two sums are equal and it collapses to $2\mathbf{Ax}$. Every quadratic form in
this book has a symmetric $\mathbf{A}$ — they are all covariance matrices or crossproducts — so
the short version is the one to carry.
:::


## Variances of linear combinations: building V

Almost everything this book computes is a variance of something built out of something else, so
this is the most-used result in it.

### From one number to many

For a scalar, multiplying a random variable by a constant multiplies its variance by the square of
that constant: $\mathrm{Var}(ax) = a^2\mathrm{Var}(x)$. Doubling every breeding value quadruples
the genetic variance.

Now the same thing with a matrix. Take three animals — a sire $S$ and his two full-sib progeny,
animals 1 and 2 — with additive relationship matrix

$$
\mathbf{A} = \begin{bmatrix} 1 & 0.5 & 0.5 \\ 0.5 & 1 & 0.5 \\ 0.5 & 0.5 & 1 \end{bmatrix}
\qquad
\mathbf{G} = \mathbf{A}\sigma^2_a = 36\mathbf{A}
$$

Only the two progeny have records, and each has one, so the matrix linking records to animals is

$$
\mathbf{Z} = \begin{bmatrix} 0 & 1 & 0 \\ 0 & 0 & 1 \end{bmatrix}
$$

Two rows, one per record; three columns, one per animal in the pedigree. The sire's column is all
zeros because he has no record of his own.

What is the covariance matrix of the two animals' breeding values *as they appear in the records*,
$\mathrm{Var}(\mathbf{Zu})$? Work it out:

$$
\mathbf{ZGZ}' =
\begin{bmatrix} 0 & 1 & 0 \\ 0 & 0 & 1 \end{bmatrix}
\begin{bmatrix} 36 & 18 & 18 \\ 18 & 36 & 18 \\ 18 & 18 & 36 \end{bmatrix}
\begin{bmatrix} 0 & 0 \\ 1 & 0 \\ 0 & 1 \end{bmatrix}
= \begin{bmatrix} 36 & 18 \\ 18 & 36 \end{bmatrix}
$$

<!-- {{< animation ch01-var-linear caption="Z, G and Z' close in on V." >}} -->

$\mathbf{Z}$ **selected**: pre-multiplying picked out the rows of $\mathbf{G}$ belonging to
recorded animals, and post-multiplying by $\mathbf{Z}'$ picked out the matching columns. What
survives is the sibs' 2 × 2 corner of $\mathbf{G}$ — the sire's row and column are gone, and the
18 off the diagonal is $0.5 \times 36$, the genetic covariance between full sibs.

**What mattered and what did not.** The 36 was $\sigma^2_a$ and the 0.5 was the sib relationship —
change either and every number changes. The dimensions came from two records and three animals.
But the *arrangement* — pre-multiply by the matrix, post-multiply by its transpose — is what you
would do for any $\mathbf{Z}$ at all, and it is the matrix version of squaring the constant.

::: {.callout-important .key-equation}
## Key Equation — Variance of a linear combination

$$
\mathrm{Var}(\mathbf{Ay}) = \mathbf{A}\,\mathrm{Var}(\mathbf{y})\,\mathbf{A}'
$$ {#eq-varlin}

**In words:** to get the variance of something built by multiplying, put the multiplier on the
left, the original variance in the middle, and the transposed multiplier on the right. The
transpose on the right is what keeps the answer square and symmetric.

Here $\mathbf{A}$ is **any matrix of constants** — this is the one place in the book where that
letter is not the relationship matrix. In the worked example above the multiplier was
$\mathbf{Z}$; below it is $(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'$.
:::

### Assembling V

Now add the residuals. Each record has residual variance $\sigma^2_e = 64$, and residuals are
independent of each other and of the breeding values, so
$\mathbf{R} = \mathbf{I}\sigma^2_e$ and the two variances simply add:

$$
\mathbf{V} = \mathrm{Var}(\mathbf{y}) = \mathbf{ZGZ}' + \mathbf{R}
= \begin{bmatrix} 36 & 18 \\ 18 & 36 \end{bmatrix}
+ \begin{bmatrix} 64 & 0 \\ 0 & 64 \end{bmatrix}
= \begin{bmatrix} 100 & 18 \\ 18 & 100 \end{bmatrix}
$$

Every number in $\mathbf{V}$ is something a breeder already knows. The diagonal is the phenotypic
variance, 100 kg², so $h^2 = 36/100 = 0.36$ — Example B's trait, unchanged. The off-diagonal 18 is
the phenotypic covariance between two full sibs, and $18/100 = 0.18$ is their phenotypic
correlation, which is $\tfrac{1}{2}h^2$ exactly as sib-analysis theory says it should be.

**That is the entire argument of the mixed model, in one matrix.** Relationships live in
$\mathbf{A}$; scaling by $\sigma^2_a$ makes them a covariance; $\mathbf{Z}$ propagates them from
animals to records; adding $\mathbf{R}$ produces the covariance structure of the data. Chapter 6
does exactly this and then solves.

### The variance of an estimate

The same rule gives standard errors. Since $\hat{\mathbf{b}} =
(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{y}$ is a linear combination of $\mathbf{y}$,
applying @eq-varlin with $\mathrm{Var}(\mathbf{y}) = \mathbf{I}\sigma^2_e$ and simplifying gives

$$
\mathrm{Var}(\hat{\mathbf{b}}) = (\mathbf{X}'\mathbf{X})^{-1}\sigma^2_e
$$

For a contrast $\mathbf{k}'\hat{\mathbf{b}}$, one more application gives
$\mathrm{Var}(\mathbf{k}'\hat{\mathbf{b}}) =
\mathbf{k}'(\mathbf{X}'\mathbf{X})^{-1}\mathbf{k}\,\sigma^2_e$. For Example A's flock difference,
$\mathbf{k} = (0, 1)'$ picks out the bottom-right element of $(\mathbf{X}'\mathbf{X})^{-1}$, which
is 1, so the variance is $1 \times 8 = 8$ and the standard error is $\sqrt{8} = 2.83$ kg. The
7 kg flock advantage is about 2.5 standard errors from zero — with four lambs, suggestive and no
more.

In Chapter 6 the same diagonal, taken from the inverted mixed model coefficient matrix instead,
becomes prediction error variance and then accuracy. It is the same calculation throughout.

```{r}
#| label: ch01-var-linear
# A is now the 3 x 3 relationship matrix: sire S, and his two full-sib progeny.
A <- matrix(c(1.0, 0.5, 0.5,
              0.5, 1.0, 0.5,
              0.5, 0.5, 1.0), nrow = 3, byrow = TRUE)

# Z links the two records to their animals. The sire's column is all zeros:
# he has no record of his own.
Z <- matrix(c(0, 1, 0,
              0, 0, 1), nrow = 2, byrow = TRUE)

sigma2_a <- 36
sigma2_e <- 64

G <- A * sigma2_a          # relationships become a covariance matrix
R <- diag(sigma2_e, 2)     # residuals, independent

ZGZt <- Z %*% G %*% t(Z)   # Var(Zu): the sibs' corner of G
ZGZt

V <- ZGZt + R              # Var(y)
V

# Everything a breeder reads off V.
sigma2_a / V[1, 1]         # h2 = 0.36
V[1, 2] / V[1, 1]          # 0.18, the sibs' phenotypic correlation, = h2 / 2
```

The simulation is the part a student actually believes. Draw 50,000 sets of breeding values,
add residuals, and the realised covariance of the records converges on $\mathbf{V}$:

```{r}
#| label: ch01-var-linear-sim
set.seed(2024)
n_sim <- 50000

L <- t(chol(G))                                    # section 1.12
u <- L %*% matrix(rnorm(3 * n_sim), nrow = 3)      # breeding values for S, 1, 2
e <- matrix(rnorm(2 * n_sim, sd = sqrt(sigma2_e)), nrow = 2)
y_sim <- Z %*% u + e

round(cov(t(y_sim)), 1)     # converges on V: 100, 18 / 18, 100
```

And the same rule gives the standard error of the flock contrast from Example A:

```{r}
#| label: ch01-contrast-se
k <- c(0, 1)                       # picks out the flock A advantage
XtX_inv <- solve(crossprod(X))
sigma2_e_hat <- 8                  # from section 1.14

var_contrast <- drop(t(k) %*% XtX_inv %*% k) * sigma2_e_hat
var_contrast
sqrt(var_contrast)                 # standard error, 2.83 kg
```

## Which section do you need for which chapter?

Nobody reads a matrix algebra chapter straight through and remembers it. Come back here.

| Section | What it gives you | Needed by |
|---|---|---|
| 1.1 Why linear algebra | what it is, and where a breeder meets it | orientation only |
| 1.2 Matrices, vectors, scalars | dimensions, elements, notation, linear combination | all |
| 1.3 Transpose | $\mathbf{A}'$; symmetry | all |
| 1.4 Addition and scalars | $\mathbf{P} = \mathbf{G} + \mathbf{R}$; $\mathbf{G} = \mathbf{A}\sigma^2_a$; $\alpha$ | Ch 3, Ch 6, Ch 11 |
| 1.5 Multiplication | the row-by-column walk; $(\mathbf{AB})' = \mathbf{B}'\mathbf{A}'$ | Ch 2, Ch 6 — the MME itself |
| 1.6 Special matrices | $\mathbf{I}$, diagonal, symmetric, triangular, block | all |
| 1.7 One model, many animals | $\mathbf{y} = \mathbf{Xb} + \mathbf{e}$; $\mathbf{X}'\mathbf{X}$ as counts, $\mathbf{X}'\mathbf{y}$ as sums | Ch 2, and every chapter after |
| 1.8 Rank and determinant | why breeding models are singular | Ch 2, Ch 5, Ch 21 |
| 1.9 Inversion and partitioning | $\mathbf{A}^{-1}$; Schur complement; do not invert to solve | Ch 4, Ch 6, Ch 7, Ch 24 |
| 1.10 Generalized inverses | non-unique solutions; estimable functions | Ch 2, Ch 9 |
| 1.11 Kronecker and Hadamard | $\mathbf{G}_0 \otimes \mathbf{A}$; $\mathbf{A} \# \mathbf{A}$ | Ch 11, Ch 16, Ch 17 |
| 1.12 Cholesky | factorising, simulating, log-determinants | Ch 4, Ch 21, Ch 22, Ch 24 |
| 1.13 Eigenvalues | positive definiteness; principal components | Ch 12, Ch 21 |
| 1.14 Quadratic forms and traces | sums of squares; degrees of freedom; derivatives | Ch 2, Ch 20, Ch 21 |
| 1.15 Variance of a linear combination | $\mathbf{V} = \mathbf{ZGZ}' + \mathbf{R}$; PEV | Ch 3, Ch 6, Ch 11, Ch 23 |

**Sections 1.2 to 1.5 are the ones to read in order.** They build on each other — multiplication
needs the transpose, and everything after §1.5 needs multiplication. From §1.6 onward you can
enter wherever you need to.

If you are heading straight for the animal model in Chapter 6, the minimum is §1.5, §1.9 and
§1.15 — but read §1.2 through §1.5 first if you have not done linear algebra before.

## Further reading

Four sources, and when to reach for each.

**[@searle1982matrix], *Matrix Algebra Useful for Statistics*.** If you read one book alongside
this chapter, read this one. Searle wrote it for statisticians rather than mathematicians, it
covers every operation here in far more depth, and its treatment of rank, generalized inverses and
quadratic forms is the natural next step from §1.8, §1.10 and §1.14. Searle took his doctorate in
animal breeding under Henderson, which is part of why the examples feel familiar.

**[@penrose1955generalized] and [@rao1962note].** The two short papers behind §1.10 — Penrose
defines the generalized inverse, Rao shows what it buys you when the normal equations are
singular. Read them in that order, and only if you want to know where `MASS::ginv()` comes from.

**[@henderson1976simple].** The paper that made the animal model practical, by giving rules to
write $\mathbf{A}^{-1}$ directly from a pedigree without forming $\mathbf{A}$. You do not need it
yet — Chapter 4 is built on it.

**[@wells2009conversation].** An interview rather than a result, and the most readable thing on
this list: how matrix algebra, linear models and animal breeding grew up together at Cornell, told
by someone who was there.

## Key equations

| Name | Equation | Section |
|---|---|---|
| The reversal rules | @eq-reversal | §1.5 |
| The inverse of a 2 × 2 | @eq-inverse2 | §1.9 |
| Variance of a linear combination | @eq-varlin | §1.15 |

## Exercises

Solutions are in Appendix H.

1. Five ewes are weighed. Three are in flock 1, with weights 120, 124 and 131 kg; two are in
   flock 2, with weights 108 and 112 kg. Using a baseline plus a flock 1 indicator, write
   $\mathbf{X}$ and $\mathbf{y}$, then form $\mathbf{X}'\mathbf{X}$ and $\mathbf{X}'\mathbf{y}$ by
   hand. State in words what every element of each counts or sums. *(objective 1)*

2. For two traits, $\mathbf{G} = \begin{bmatrix} 16 & 6 \\ 6 & 9 \end{bmatrix}$ and
   $\mathbf{R} = \begin{bmatrix} 48 & 2 \\ 2 & 16 \end{bmatrix}$. Form $\mathbf{P}$, then compute
   both heritabilities, the genetic correlation, and the phenotypic correlation. Which of the two
   traits would respond faster to selection, and why is that not the whole answer?
   *(objective 1)*

3. Take the $\mathbf{X}'\mathbf{X}$ from exercise 1. Compute its determinant, invert it by hand
   using @eq-inverse2, and solve for $\hat{\mathbf{b}}$. Verify with `solve()`, and confirm your
   two elements against the raw flock means. *(objective 1)*

4. Six lambs in three flocks of two weigh 50 and 54; 44 and 48; 39 and 41 kg. Build the
   overparameterised $\mathbf{X}$ with a baseline and all three flock columns. Show it is rank
   deficient and give its rank. Then produce two different generalized-inverse solutions, show
   that the fitted values agree, and give one estimable function on which they agree.
   *(objective 2)*

5. With $\mathbf{G}_0 = \begin{bmatrix} 36 & 12 \\ 12 & 9 \end{bmatrix}$ for two traits and
   $\mathbf{A}$ the relationship matrix of three animals — a sire and two of his half-sib progeny
   out of unrelated dams — form both $\mathbf{G}_0 \otimes \mathbf{A}$ and
   $\mathbf{A} \otimes \mathbf{G}_0$. In each, identify the element that is
   $\mathrm{cov}(a_{1,\,\text{trait }1},\, a_{3,\,\text{trait }2})$, and say which ordering each
   matrix uses. *(objective 3)*

6. Let $\mathbf{G} = \begin{bmatrix} 64 & 24 \\ 24 & 25 \end{bmatrix}$ be a genetic covariance
   matrix for two traits.
   a. Compute its Cholesky factor $\mathbf{L}$ by hand, entry by entry, and verify
      $\mathbf{LL}' = \mathbf{G}$. *(objective 4)*
   b. In R, use $\mathbf{L}$ to simulate 10,000 pairs of correlated breeding values and confirm
      that the realised covariance matrix recovers $\mathbf{G}$. Remember that `chol()` returns
      the upper factor. *(objective 4)*
   c. Those 10,000 animals are each measured once for trait 1, so
      $\mathbf{Z} = \mathbf{I}$ and $\mathbf{R} = \mathbf{I}\sigma^2_e$ with
      $\sigma^2_e = 36$. Use @eq-varlin to give $\mathrm{Var}(\mathbf{Zu})$ and then
      $\mathbf{V}$ for trait 1, and state the heritability. *(objective 5)*
