---
title: "CHAPTER TITLE"
---

<!--
  Chapter skeleton. Copy to chapters/NN-slug.qmd and replace everything in CAPS.
  The standards this implements: plans/chapter_standards.md (structure, callouts)
                                 plans/pedagogical_order.md (order of presentation)
  Every callout below appears exactly once as a live example. Delete the ones
  this chapter does not need — the required ones are marked REQUIRED.
  Budget: 15 boxes maximum per chapter.

  THE ORDER OF THE SECTIONS BELOW IS THE STANDARD. Numbers before symbols:
  the reader meets every object as literal numbers, solves with them, and only
  then is given the general form. Do not move "The general form" earlier, and
  do not let a Key Equation box be the first sight of its own content.
-->

::: {.callout-note icon=false}
## Learning Objectives

By the end of this chapter, you will be able to:

1. **VERB** the **OBJECT** — copied verbatim from `CHAPTERS.md`
2. **VERB** the **OBJECT**
3. **VERB** the **OBJECT**

<!-- 3-5 items. Verb first. Never "understand" or "be familiar with" —
     if it can't be tested by an exercise in this chapter, it isn't an objective. -->
:::

*Assumes Chapter N (**SYMBOL**) and Chapter N (**SYMBOL**).*

## Why WHAT THIS CHAPTER DOES

<!-- The problem, in breeding terms, before any algebra. A reader who stops
     after this section should still know what question the chapter answers. -->

<!-- Chapter-opener clip, if this chapter has one:
{{< animation chNN-opener caption="ONE SENTENCE" >}} -->

::: {.callout-note .practice}
## In Practice — WHERE

WHAT A REAL EVALUATION DOES WITH THIS. Name the organisation, the species, the scale.
:::

## The data

<!-- 5-10 animals. Species named. Read from data/, never inlined (CLAUDE.md). -->

```{r}
#| label: load-data
#| message: false
library(Matrix)
dat <- read.csv(here::here("data", "chNN-DATASET.csv"))
knitr::kable(dat)
```

TWO SENTENCES ON WHY THIS TRAIT AND THIS SPECIES. Check the reserved-trait table in
`CLAUDE.md` before choosing.

## The matrices for these ANIMALS

<!-- BEAT 6. Literal numbers only. No general dimensions, no Tier 1 boxes, no
     symbol that is not a name for an array already printed on this page.
     Wrong: "Let X be the n x p incidence matrix of fixed effects."
     Right: print the array, say what each row and column is, then: "Call this X." -->

Every animal is in the same contemporary group and is either a barrow or a gilt, so each record
needs two indicators:

$$
\begin{bmatrix} 1 & 0 \\ 1 & 0 \\ 1 & 1 \\ \vdots & \vdots \end{bmatrix}
$$

N rows, one per record; two columns, one per thing being estimated. Call this $\mathbf{X}$.

<!-- Repeat for Z, A-inverse, alpha. Same discipline each time. -->

::: {.callout-note .definition}
## Definition — TERM

WHAT IT IS, in one or two plain sentences. The formal condition second, if it differs.

**Why it matters.** What goes wrong if you do not know this, and the chapter where it returns.

<!-- Goes immediately AFTER the term has been met in context, never before.
     Outside the 15-box cap. Only for vocabulary the reader is held to later —
     a symbol defined on first use is prose, not a box. -->
:::

::: {.callout-note .notation}
## Notation Watch — SYMBOL

Some texts write X where this book writes Y. WHY THE DIFFERENCE EXISTS. Appendix A is
authoritative for this book.

<!-- Only for genuine conflicts in the literature. Defining a symbol on first
     use is prose, not a box. -->
:::

## Solving by hand

<!-- BEAT 7. Tier 2 arithmetic, every step shown. Board-work clips go at the
     hard steps — where the difficulty is a SEQUENCE the page cannot carry.
     Where a clip exists, it is the instance and the prose does not re-narrate
     the steps. -->

<!-- {{< animation chNN-STEP caption="ONE SENTENCE" >}} -->

::: {.callout-warning}
## Common Mistake — WHAT

THE ERROR STUDENTS ACTUALLY MAKE, what it looks like when it happens, and how to catch it.
:::

::: {.callout-caution .check collapse="true"}
## Check Yourself

<!-- Sits HERE: after the worked instance, before the general form. The one
     point in the chapter where a prediction is possible and not yet spoiled. -->

You solved this for N animals and got an M x M system. Before you read on: for 600 animals in 12
contemporary groups, how large is the system, and which block grew?

**Answer.** THE ANSWER, plus the one-line reason it is the answer.
:::

## The general form

<!-- BEAT 8. The beat that gets dropped under deadline. Do not drop it: without
     it the reader has learned an instance, not a model. -->

<!-- THE PATTERN PARAGRAPH. Three questions, in this order, in 3-8 lines.
     Delete these comments, not the paragraph.
       1. What in that calculation depended on the particular numbers?
       2. What depended on the shape of the data?
       3. What would have been identical for ANY dataset?
     Question 3's answer, written out, IS the Key Equation below. -->

The WEIGHTS were these animals' — change them and the solutions change. The size of the system was
set by NUMBER OF ANIMALS and NUMBER OF GROUPS. But the *arrangement* — WHAT — would have been
identical for six animals or six million.

::: {.callout-important .key-equation}
## Key Equation — NAME OF THE RESULT {#eq-LABEL}

$$
\text{THE EQUATION}
$$ {#eq-LABEL-math}

**In words:** ONE SENTENCE SAYING WHAT IT DOES, with no symbols in it.

<!-- Tier 1. 1-3 per chapter. An equation label means Tier 1 and nothing else
     in the chapter carries one. CHECK BEFORE SHIPPING: is every symbol in this
     box one the reader has already seen carrying a number? If not, the chapter
     is written backwards. -->
:::

::: {.callout-note .derivation collapse="true"}
## Derivation — WHAT IS DERIVED (optional)

Skip this on a first read. You already have the result above; this only shows that it holds for
every dataset and not just the one you solved.

THE DERIVATION. If it runs past about a page, put it in an appendix and leave a
four-line sketch here with a cross-reference.

<!-- Tier 3, and it goes AFTER the box, not before. -->
:::

## What the answer means

<!-- BEAT 9. The other beat most often dropped, and the one students most need.
     What would a breeder do differently because of these numbers? Which animal
     do you keep? -->

## The same model in R

```{r}
#| label: by-hand-in-r
# Explicit matrix code first: build every matrix, no shortcuts.
```

```{r}
#| label: with-a-package
# Then the one-line package call. It must reproduce the hand answer exactly.
```

The hand solution and the package solution agree to N decimal places.

::: {.callout-tip}
## In R — WHAT

THE R-SPECIFIC TRAP OR IDIOM. Keep it about R, not about the model.
:::

## Scaling up

<!-- Same trait, same model, ~2,000 animals. Show that nothing conceptual changed. -->

::: {.callout-note .beyond}
## Beyond This Book — TOPIC

One or two sentences on the genomic or large-scale successor to what this chapter
teaches, and where it is covered. Do not teach it here.
:::

## Key equations

| | Equation | Where |
|---|---|---|
| NAME | @eq-LABEL | §SECTION |

<!-- Every Tier 1 box in this chapter, one line each. These also feed the
     key-equation index in Appendix A. -->

## Exercises

<!-- Count per CHAPTERS.md. Every learning objective is tested by at least one.
     Solutions go to appendices/H-solutions.qmd, not here. -->

1. EXERCISE
2. EXERCISE
